---
title: "Compile Decoy Intron Data (overlap first, then bash features)"
output: html_document
---

<!--
  Four-part workflow, with a bash step after each of the first three:
  Part 1: Load cryptic_sites.bed (deep-intronic SpliceAI sites with no RBPnet/PRPF8/SmB support at the site), overlap with Vast-DB PSI Table introns, find each intron's canonical 5' splice site in Canonical_splice_sites.bed, and write those windows with the cryptic ID in column 4.
  After Part 1: Intersect the windows with the RBPnet, PRPF8 and SmB peaks (intersect_spliceai_support.sh) -> canonical_supported_sites.bed.
  Part 2: Keep the cryptic sites whose canonical 5' splice site is supported -> cryptics_supported.bed.
  After Part 2: Re-score decoys.bed and cryptics_supported.bed with SpliceAI_Inference.py (local 49 nt donor score).
  Part 3: Load both re-scored BEDs as one table (site_class = decoy / cryptic_supported), filter to local SpliceAI >= 0.1, remove annotated alternative 5' splice sites (Vast-DB HsaALTD donors) and write the final BEDs, overlap with PSI Table introns, extract strand-aware distance from the canonical 5' splice site, write step1 TSV + BEDs for bash processing.
  After Part 3: Run extract_phastcons_scores.sh to calculate average phastCons scores over intron segments harboring each site. Use run_maxentscan_canonical.sh and run_maxentscan_decoy.sh to calculate MaxEntScan score for both the site and its canonical 5' splice site in bash. Run extract_gc_content.sh for intron and 49 nt site GC. These run once on the combined BEDs, so decoys and cryptics get the same features.
  Part 4: Reload step1 table, merge scores, plot, write the final feature table (decoys and supported cryptics, same columns) and a per-class summary.

  Every splice-site BED here is 10 columns: BED6 + rbpnet/prpf8/smb support
  counts + n_categories. After the re-score by scripts/SpliceAI_Inference.py,
  column 5 is the LOCAL donor score.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-libraries}
library(data.table)
library(GenomicRanges)
library(ggplot2)
```

```{r paths-v3}
# Upstream of this document (see README "Ordering (v3)"):
#   decoy_exon_overlaps.Rmd Part 1  protein_coding keep -> spliceai_05min_proteincoding.bed
#   split_canonical_exonic.sh       canonical split     -> deep_intronic_splice_sites.bed
#   decoy_exon_overlaps.Rmd Part 2  exon removal        -> deep_intronic_noexon_splice_sites.bed
#   intersect_spliceai_support.sh   CLIP support        -> decoys.bed / cryptic_sites.bed
out_dir <- "../results"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
psi_table <- "../reference/PSI_TABLE-hg38.tab.gz"

# Parts 1-2: supported-cryptic selection.
cryptic_sites_bed <- file.path(out_dir, "cryptic_sites.bed")
canonical_sites_bed <- "../data/Decoys/Canonical_splice_sites.bed"
cryptic_canonical_bed <- file.path(out_dir, "cryptic_canonical_sites.bed")
canonical_supported_bed <- file.path(out_dir, "canonical_supported_sites.bed")
cryptics_supported_bed <- file.path(out_dir, "cryptics_supported.bed")

# Part 3: both classes after the local SpliceAI re-score. The names become the
# site_class column.
splicescores_beds <- c(
  decoy = "../data/Decoys/decoys_splicescores.bed",
  cryptic_supported = "../data/Decoys/cryptics_supported_splicescores.bed"
)

# Minimum LOCAL SpliceAI donor score to retain (applied in Part 3, after load).
# The re-score replaces the whole-intron SpliceAI inference with a LOCAL donor
# score over a 49 nt window centred on each site, so this is NOT comparable to
# the >= 0.05 threshold applied upstream on the raw inference set.
MIN_LOCAL_SPLICEAI <- 0.1

step1_tsv <- file.path(out_dir, "decoy_intron_overlap_step1.tsv")

# Shared by decoys.bed, cryptic_sites.bed, cryptics_supported.bed and their
# re-scored versions.
site_cols <- c(
  "chr", "start", "end", "name", "SpliceAI_score", "strand",
  "rbpnet", "prpf8", "smb", "n_categories"
)
```

## PSI table: HsaIN only, drop `-Q` columns, parse intron coordinates

Loaded once, for Part 1 (cryptic → canonical donor) and Part 3 (intron features).

```{r load-psi-v3}
psi <- fread(psi_table, sep = "\t", header = TRUE)

# Annotated alternative 5' splice sites, removed in Part 3. FullCO lists every
# donor of an Alt5 event: + strand "chr:acceptor-D1+D2,next", - strand
# "chr:D1+D2-acceptor,next".
altd_co <- unique(psi[grepl("^HsaALTD", EVENT), FullCO])
altd_donors <- rbindlist(lapply(strsplit(altd_co, ":", fixed = TRUE), function(x) {
  ends <- strsplit(sub(",.*", "", x[2]), "-", fixed = TRUE)[[1]]
  minus <- grepl("+", ends[1], fixed = TRUE)
  d <- strsplit(if (minus) ends[1] else ends[2], "+", fixed = TRUE)[[1]]
  data.table(chr = x[1], donor = suppressWarnings(as.integer(d)), strand = if (minus) "-" else "+")
}))
altd_donors <- unique(altd_donors[!is.na(donor)])
cat("Annotated alternative 5' splice sites (HsaALTD donors):", nrow(altd_donors), "\n")

psi <- psi[grepl("^HsaIN", EVENT)]
q_cols <- grep("-Q$", names(psi), value = TRUE)
if (length(q_cols) > 0) psi[, (q_cols) := NULL]

coord_parts <- tstrsplit(psi$COORD, ":", fixed = TRUE)
range_parts <- tstrsplit(coord_parts[[2]], "-", fixed = TRUE)
psi[, chr := coord_parts[[1]]]
psi[, intron_start := as.integer(range_parts[[1]])]
psi[, intron_end := as.integer(range_parts[[2]])]
psi <- psi[!is.na(chr) & !is.na(intron_start) & !is.na(intron_end)]
cat("PSI rows after filters:", nrow(psi), "\n")

# The PSI table carries no strand; each site's own strand orients its intron.
psi_gr <- GRanges(
  seqnames = psi$chr,
  ranges = IRanges(start = psi$intron_start, end = psi$intron_end)
)
```

# Part 1 — Match cryptic sites to their canonical 5' splice site

`cryptic_sites.bed` holds deep-intronic SpliceAI sites with no RBPnet, PRPF8 or SmB support at the site itself. They are kept only if the canonical 5' splice site of their intron is supported by RBPnet, PRPF8 or SmB. This part finds that donor. Each cryptic site is overlapped with the Vast-DB introns. Each intron's donor (`+`: intron start, `−`: intron end) must fall in a `Canonical_splice_sites.bed` window, which spans ±5 nt around an annotated canonical donor. The matched windows are written with the cryptic ID (`GENE_start`) in column 4, ready for the bedtools intersect.

```{r load-cryptics}
if (!file.exists(cryptic_sites_bed)) {
  stop(
    "Cryptic sites not found: ", cryptic_sites_bed, "\n",
    "  It is the unsupported branch of the step-3 CLIP intersect:\n",
    "    bash scripts/intersect_spliceai_support.sh -w 0 \\\n",
    "      -a data/Decoys/deep_intronic_noexon_splice_sites.bed -o results/clip_support \\\n",
    "      RBPNET=... PRPF8=... SmB=...\n",
    "    mv results/clip_support_w0_nohits.bed results/cryptic_sites.bed"
  )
}
cryptic <- fread(cryptic_sites_bed, sep = "\t", header = FALSE, col.names = site_cols)
cryptic[, decoyID := paste0(name, "_", start)]
cat("Cryptic sites:", nrow(cryptic), " unique IDs:", uniqueN(cryptic$decoyID), "\n")
```

```{r match-cryptic-canonical}
canon <- fread(
  canonical_sites_bed,
  sep = "\t", header = FALSE,
  col.names = c("chr", "start", "end", "gene", "score", "strand")
)
# An earlier build of this file sat 400 nt into the flanking exon and covered
# no real donor. The HNRNPH1 intron-4 donor catches that.
if (nrow(canon[chr == "chr5" & strand == "-" & start < 179620891L & end >= 179620891L]) == 0) {
  stop(
    "No Canonical_splice_sites.bed window covers the HNRNPH1 intron-4 donor ",
    "(chr5:179,620,891). Regenerate it with scripts/CreateCanonicalSpliceSiteBed.sh"
  )
}
canon_gr <- GRanges(canon$chr, IRanges(canon$start + 1L, canon$end), strand = canon$strand)

cryptic_gr <- GRanges(
  cryptic$chr, IRanges(cryptic$start + 1L, cryptic$end), strand = cryptic$strand
)
ci <- findOverlaps(cryptic_gr, psi_gr, ignore.strand = TRUE)

# One donor per (site, intron). Same strand logic as canonical_coords_for_maxent.bed in Part 3.
pairs <- unique(data.table(
  q = queryHits(ci),
  chr = cryptic$chr[queryHits(ci)],
  strand = cryptic$strand[queryHits(ci)],
  donor = ifelse(
    cryptic$strand[queryHits(ci)] == "+",
    psi$intron_start[subjectHits(ci)],
    psi$intron_end[subjectHits(ci)]
  )
))
donor_gr <- GRanges(pairs$chr, IRanges(pairs$donor, width = 1L), strand = pairs$strand)
dc <- findOverlaps(donor_gr, canon_gr)

# A site in overlapping introns can match several canonical donors; keep them
# all so the site passes if any of them is bound.
matched <- unique(data.table(q = pairs$q[queryHits(dc)], w = subjectHits(dc)))
cryptic_canonical <- data.table(
  chr = canon$chr[matched$w],
  start = canon$start[matched$w],
  end = canon$end[matched$w],
  decoyID = cryptic$decoyID[matched$q],
  SpliceAI_score = cryptic$SpliceAI_score[matched$q],
  strand = canon$strand[matched$w]
)
setorder(cryptic_canonical, chr, start, decoyID)

n_cryptic <- nrow(cryptic)
n_in_intron <- uniqueN(queryHits(ci))
n_canon <- uniqueN(matched$q)
cat("  Cryptic sites:                              ", n_cryptic, "\n")
cat("  ... inside a Vast-DB intron:                ", n_in_intron, "\n")
cat("  ... with a canonical donor for that intron: ", n_canon, "\n")
cat("  Dropped, no Vast-DB intron:                 ", n_cryptic - n_in_intron, "\n")
cat("  Dropped, no intron with a canonical donor:  ", n_in_intron - n_canon, "\n")
cat("  Canonical windows written (site x window):  ", nrow(cryptic_canonical), "\n")
```

```{r write-cryptic-canonical}
fwrite(
  cryptic_canonical, cryptic_canonical_bed,
  sep = "\t", col.names = FALSE, quote = FALSE, eol = "\n"
)
cat("Wrote", cryptic_canonical_bed, " rows:", nrow(cryptic_canonical), "\n")
```

---

## Between Part 1 and Part 2 — canonical 5'SS support intersect

#### **Stop here after knitting Part 1** and run from `hnRNPH1_IR_MAF`:

#### `$ bash scripts/intersect_spliceai_support.sh -a results/cryptic_canonical_sites.bed -o results/canonical_support -w 0 RBPNET=data/Decoys/rbpnet_clippy_f50_rollmean10_minHeightAdjust1.0_minPromAdjust1.0_minGeneCount5_Peaks.bed PRPF8=data/CLIP/PRPF8_clippy_w0.5_rollmean80_minHeightAdjust3.0_minPromAdjust3.0_minGeneCount5_Peaks.bed SmB=data/CLIP/SmB_clippy_n40_w0.5_rollmean40_minHeightAdjust8.0_minPromAdjust5.0_minGeneCount5_Peaks.bed`
#### `$ mv results/canonical_support_w0.bed results/canonical_supported_sites.bed`

#### The same three tracks as the step-3 site intersect count as support. RBPnet hits 134,621 of the 192,965 canonical windows (70%) and PRPF8 or SmB 78,626 (41%), so most of the selectivity comes from the canonical-donor match itself rather than from this intersect.

#### The output is BED6 + `RBPNET` + `PRPF8` + `SmB` counts + `hits`, and holds only the supported windows (`hits > 0`).

# Part 2 — Keep cryptic sites whose canonical 5' splice site is bound

```{r filter-cryptics-supported}
if (!file.exists(canonical_supported_bed)) {
  stop(
    "Canonical support not found: ", canonical_supported_bed, "\n",
    "  Run the intersect in 'Between Part 1 and Part 2', then knit again."
  )
}
canon_sup <- fread(canonical_supported_bed, sep = "\t", header = FALSE)
if (ncol(canon_sup) != 10) {
  stop(
    "Expected 10 columns (BED6 + RBPNET + PRPF8 + SmB + hits), got ", ncol(canon_sup),
    ". Was the intersect run with exactly RBPNET, PRPF8 and SmB, in that order?"
  )
}
setnames(canon_sup, c(
  "chr", "start", "end", "decoyID", "SpliceAI_score", "strand",
  "canon_rbpnet", "canon_prpf8", "canon_smb", "canon_hits"
))

# Every ID must come from this knit's Part 1. Otherwise the intersect output
# was built from an earlier cryptic_sites.bed and would filter the wrong sites.
stale <- setdiff(canon_sup$decoyID, cryptic_canonical$decoyID)
if (length(stale) > 0) {
  stop(
    length(stale), " IDs in ", canonical_supported_bed,
    " are not in this run's Part 1 output; re-run the intersect."
  )
}

per_site <- canon_sup[canon_hits > 0, .(
  canon_rbpnet = any(canon_rbpnet > 0),
  canon_prpf8 = any(canon_prpf8 > 0),
  canon_smb = any(canon_smb > 0)
), by = decoyID]

cryptics_supported <- cryptic[decoyID %in% per_site$decoyID]
fwrite(
  cryptics_supported[, ..site_cols], cryptics_supported_bed,
  sep = "\t", col.names = FALSE, quote = FALSE, eol = "\n"
)

cat("  Cryptic sites with a canonical donor:       ", n_canon, "\n")
cat("  ... donor supported by RBPnet, PRPF8 or SmB (kept):", nrow(cryptics_supported), "\n")
cat("      by track:  RBPnet", per_site[canon_rbpnet == TRUE, .N],
    " PRPF8", per_site[canon_prpf8 == TRUE, .N],
    " SmB", per_site[canon_smb == TRUE, .N], "\n")
cat("      RBPnet only (would be lost without it):", per_site[canon_rbpnet & !canon_prpf8 & !canon_smb, .N], "\n")
cat("Wrote", cryptics_supported_bed, "\n")

decoys_bed <- file.path(out_dir, "decoys.bed")
n_decoys <- if (file.exists(decoys_bed)) length(readLines(decoys_bed)) else NA_integer_
cat(sprintf(
  "Re-score estimate at ~0.65 s/site: cryptics_supported %.1f h, decoys %.1f h\n",
  nrow(cryptics_supported) * 0.65 / 3600, n_decoys * 0.65 / 3600
))
```

---

## Between Part 2 and Part 3 — local SpliceAI re-score

#### **Stop here after knitting Part 2** and re-score both classes from `hnRNPH1_IR_MAF`. Each run checkpoints to `<out>.scores`; add `--resume` to continue an interrupted run.

#### `$ conda activate spliceai-env`
#### `$ python3 scripts/SpliceAI_Inference.py --bed results/decoys.bed --fasta ../../reference/genomes/Gencode49/GRCh38.primary_assembly.genome.fa --genome ../../reference/genomes/Gencode49/genome.sizes --batch-size 64 --out data/Decoys/decoys_splicescores.bed`
#### `$ python3 scripts/SpliceAI_Inference.py --bed results/cryptics_supported.bed --fasta ../../reference/genomes/Gencode49/GRCh38.primary_assembly.genome.fa --genome ../../reference/genomes/Gencode49/genome.sizes --batch-size 64 --out data/Decoys/cryptics_supported_splicescores.bed`

# Part 3 — Load re-scored decoys and supported cryptics, overlap introns, write intermediate table

## Load re-scored BEDs and filter on local SpliceAI score

```{r load-splice}
missing_scores <- splicescores_beds[!file.exists(splicescores_beds)]
if (length(missing_scores) > 0) {
  stop(
    "Re-scored BED(s) not found:\n  ", paste(missing_scores, collapse = "\n  "), "\n",
    "  Run the re-score in 'Between Part 2 and Part 3', then knit again."
  )
}

# 10-column BEDs. SpliceAI_Inference.py rewrites only column 5 (the score) and
# passes every other column through, so the support counts from the intersect
# survive the re-score and stay available downstream.
bed <- rbindlist(lapply(names(splicescores_beds), function(cls) {
  x <- fread(splicescores_beds[[cls]], sep = "\t", header = FALSE, col.names = site_cols)
  x[, site_class := cls]
}))
setnames(bed, "name", "decoyID")
# decoyID becomes <gene>_<start>, the key used throughout this repo. Supported
# cryptics use the same key; site_class says which class a row belongs to.
bed[, decoyID := paste0(decoyID, "_", start)]

# The re-scored cryptic file must be the re-score of the CURRENT
# cryptics_supported.bed. A re-score of an earlier support definition has the
# same layout and would otherwise pass through silently.
if (file.exists(cryptics_supported_bed)) {
  cur_ids <- fread(cryptics_supported_bed, sep = "\t", header = FALSE,
                   col.names = site_cols)[, paste0(name, "_", start)]
  got_ids <- bed[site_class == "cryptic_supported", decoyID]
  if (!setequal(cur_ids, got_ids)) {
    stop(
      splicescores_beds[["cryptic_supported"]], " has ", length(got_ids),
      " sites but ", cryptics_supported_bed, " has ", length(cur_ids),
      "; re-score the current cryptics_supported.bed before knitting Part 3."
    )
  }
}

# Decoys are CLIP-supported at the site and cryptics are not, so the classes
# cannot share an ID. A shared one means one of the inputs is stale.
shared_ids <- intersect(
  bed[site_class == "decoy", decoyID],
  bed[site_class == "cryptic_supported", decoyID]
)
if (length(shared_ids) > 0) {
  stop(length(shared_ids), " IDs are both decoy and cryptic_supported, e.g. ", shared_ids[1])
}

# --- Filter on the LOCAL SpliceAI donor score -------------------------------
# The re-score computes a donor score over a 49 nt window centred on each site,
# rather than the whole-intron inference used upstream. A site can be supported
# by CLIP yet score poorly once evaluated on its own local sequence context;
# this drops those. Threshold set by MIN_LOCAL_SPLICEAI in the paths chunk.
n_before <- bed[, .(before = .N), by = site_class]
bed <- bed[SpliceAI_score >= MIN_LOCAL_SPLICEAI]
n_after <- bed[, .(after = .N), by = site_class]
cat("Local SpliceAI filter (>= ", MIN_LOCAL_SPLICEAI, "):\n", sep = "")
print(merge(n_before, n_after, by = "site_class", all.x = TRUE))
if (nrow(bed) == 0) stop("No sites passed the local SpliceAI filter.")

cat("Support categories at the site, by class:\n")
print(table(bed$site_class, bed$n_categories))

cat(
  "Rows:", nrow(bed), " Unique decoyID:", uniqueN(bed$decoyID),
  " Duplicates:", sum(duplicated(bed$decoyID)), "\n"
)
```

## Remove annotated alternative 5' splice sites

A site that sits exactly on a donor of a Vast-DB `HsaALTD` (Alt5) event is a known, used 5' splice site rather than a decoy or cryptic, so it is removed from both classes. The match is exact, on the same strand: the SpliceAI site at BED start + 1 is the donor coordinate. On `results/exonic_splice_sites.bed`, 45,088 sites match ALTD donors at offset 0, against at most 730 at ±1–2 nt. The filtered sets are written as the final BEDs.

```{r remove-altd}
altd_keys <- altd_donors[, paste(chr, donor, strand)]
bed[, is_altd := paste(chr, start + 1L, strand) %chin% altd_keys]
cat("Annotated alternative 5' splice sites removed:\n")
print(bed[, .(sites = .N, altd = sum(is_altd), kept = sum(!is_altd)), by = site_class])
bed <- bed[is_altd == FALSE][, is_altd := NULL]

final_beds <- c(
  decoy = file.path(out_dir, "decoys_final.bed"),
  cryptic_supported = file.path(out_dir, "cryptics_supported_final.bed")
)
for (cls in names(final_beds)) {
  x <- bed[site_class == cls, .(
    chr, start, end, name = sub("_[0-9]+$", "", decoyID), SpliceAI_score, strand,
    rbpnet, prpf8, smb, n_categories
  )]
  fwrite(x, final_beds[[cls]], sep = "\t", col.names = FALSE, quote = FALSE, eol = "\n")
  cat("Wrote", final_beds[[cls]], " rows:", nrow(x), "\n")
}
```

## Overlap sites with introns (shortest intron per site)

```{r overlap-v3}
decoy_gr <- GRanges(
  seqnames = bed$chr,
  ranges = IRanges(start = bed$start + 1, end = bed$end),
  strand = bed$strand
)
hits <- findOverlaps(decoy_gr, psi_gr, ignore.strand = TRUE)
cat("Overlap pairs:", length(hits), "\n")

bed_hit <- bed[queryHits(hits)]
psi_hit <- psi[subjectHits(hits)]
overlap_cols <- intersect(names(bed_hit), names(psi_hit))
if (length(overlap_cols) > 0) setnames(psi_hit, overlap_cols, paste0("psi_", overlap_cols))
merged <- cbind(bed_hit, psi_hit)

merged[, intron_length := intron_end - intron_start + 1L]
setorder(merged, decoyID, intron_length)
merged <- merged[!duplicated(decoyID)]
cat("Rows after shortest-intron selection:", nrow(merged), "\n")
```

## Decoys dropped at overlap (no matching PSI intron)
Rows in `bed` with no overlap are dropped here. This block lists those decoys and writes a small BED of **all** splice scores intervals for the **genes** they belong to, for inspection (e.g. IGV).

Gene symbol is parsed from `decoyID` as `GENE_start` by stripping the trailing `_<digits>`.

```{r overlap-dropouts-genes-bed}
# Decoys present in splicescores but absent after overlap + shortest-intron dedup
dropped_ids <- setdiff(bed$decoyID, merged$decoyID)
cat(
  "Splice Scores rows:", nrow(bed), " — kept after overlap:", nrow(merged),
  " — dropped:", length(dropped_ids), "\n"
)
if (length(dropped_ids) > 0) {
  cat("Dropped decoyIDs:\n")
  print(dropped_ids)
}

bed[, gene := sub("_[0-9]+$", "", decoyID)]
genes_dropped <- unique(bed[decoyID %in% dropped_ids, gene])

cat("Genes among dropped decoys (", length(genes_dropped), "):\n", sep = "")
print(sort(genes_dropped))

dropout_bed <- file.path(out_dir, "dropped_overlap_genes_all_splicescores_decoys.bed")
bed_invest <- bed[gene %in% genes_dropped][order(gene, chr, start)]
fwrite(
  bed_invest[, .(chr, start, end, decoyID, SpliceAI_score, strand)],
  dropout_bed,
  sep = "\t",
  col.names = FALSE,
  quote = FALSE
)
cat(
  "Wrote BED (all splice scores decoys in those genes): ", dropout_bed,
  " rows:", nrow(bed_invest), "\n",
  sep = ""
)
```

## Strand-aware distance from canonical 5' splice site

```{r distance-v3}
merged[, decoy_coord_1based := start + 1L]
merged[, decoy_distance_from_canonical_5ss := ifelse(
  strand == "+",
  decoy_coord_1based - intron_start, # decoy distance caluclated from start coord
  intron_end - decoy_coord_1based # decoy distance calculated from end coord
)]
```

## Write Part-1 output (main table for bash + Part 2)

```{r write-step1}
fwrite(merged, step1_tsv, sep = "\t", quote = FALSE)
cat("Wrote:", step1_tsv, " rows:", nrow(merged), " cols:", ncol(merged), "\n")
```

## BED files for bash (phastCons + MaxEnt)

`decoy_intron_overlap_step1.tsv` alone is not enough for external tools: **phastCons** and **MaxEntScan** run in bash on **BED intervals** (and the genome FASTA for MaxEnt). This chunk writes the three BEDs that match the overlap table—**run it whenever you knit Part 1** so the commands in the “Between” section have the expected inputs. (You only skip it if you build equivalent BEDs yourself.)

- **Intron BED** — full intron span → `extract_phastcons_scores.sh` / `bigWigAverageOverBed` (mean conservation over the intron).
- **Decoy BED** — splice scores decoy interval → `run_maxentscan_decoys.sh`.
- **Canonical BED** — strand-aware canonical 5′SS (`+`: intron start; `−`: intron end) → `run_maxentscan_canonical.sh`.

```{r write-bash-input-beds}
intron_bed <- file.path(out_dir, "intron_segments_for_phastcons.bed")
decoy_bed <- file.path(out_dir, "decoy_coords_for_maxent.bed")
canonical_bed <- file.path(out_dir, "canonical_coords_for_maxent.bed")

fwrite(
  merged[, .(chr, intron_start - 1L, intron_end, decoyID, SpliceAI_score, strand)],
  intron_bed,
  sep = "\t",
  col.names = FALSE,
  eol = "\n"
)
fwrite(
  merged[, .(chr, start, end, decoyID, SpliceAI_score, strand)],
  decoy_bed,
  sep = "\t",
  col.names = FALSE,
  eol = "\n"
)
canonical_dt <- merged[, .(
  chr = chr,
  start = ifelse(strand == "+", intron_start - 1L, intron_end - 1L),
  end = ifelse(strand == "+", intron_start, intron_end),
  decoyID = decoyID,
  SpliceAI_score = SpliceAI_score,
  strand = strand
)]
fwrite(canonical_dt, canonical_bed, sep = "\t", col.names = FALSE, eol = "\n")

# Remove any CR characters so bash/awk strand parsing always sees "+" / "-" exactly for MaxEntScan
normalize_lf <- function(path) {
  txt <- readChar(path, nchars = file.info(path)$size, useBytes = TRUE)
  txt <- gsub("\r\n", "\n", txt, fixed = TRUE)
  txt <- gsub("\r", "\n", txt, fixed = TRUE)
  writeChar(txt, path, eos = NULL, useBytes = TRUE)
}
normalize_lf(intron_bed)
normalize_lf(decoy_bed)
normalize_lf(canonical_bed)

cat("Wrote:\n ", intron_bed, "\n ", decoy_bed, "\n ", canonical_bed, "\n")
```

---

## Between Part 3 and Part 4 — bash intermediate (phastCons + MaxEnt + GC)

#### **Stop here after knitting Part 3** then run the following in the shell from `hnRNPH1_IR_MAF` (adjust paths if needed). The BEDs hold both decoys and supported cryptics, so each command runs once and both classes get the same features.

### phastCons (mean over each **intron** segment)

#### Use `scripts/extract_phastcons_scores.sh` (or direct `bigWigAverageOverBed`) with: phastcons bigwig and BED file input coordinates of intron of interest. Usage:

#### `$ bash scripts/extract_phastcons_scores.sh reference/hg38.phastCons100way.bw results/intron_segments_for_phastcons.bed > results/phastcons100_by_decoy.tsv`
#### `$ bash scripts/extract_phastcons_scores.sh reference/hg38.phastCons470way.bw results/intron_segments_for_phastcons.bed > results/phastcons470_by_decoy.tsv`

### MaxEntScan (decoy site vs canonical 5'SS)

#### Use `scripts/run_maxentscan_decoys.sh` and `scripts/run_maxentscan_canonical.sh` with genome FASTA as third argument. Usage:

#### `$ bash scripts/run_maxentscan_decoys.sh results/decoy_coords_for_maxent.bed results/maxent_decoy.bed ../../reference/genomes/Gencode49/GRCh38.primary_assembly.genome.fa`
#### `$ bash scripts/run_maxentscan_canonical.sh results/canonical_coords_for_maxent.bed results/maxent_canonical.bed ../../reference/genomes/Gencode49/GRCh38.primary_assembly.genome.fa`

#### Scores are appended as the **last column** of each output BED (column 4 = `decoyID`).

### GC content (intron and 49 nt site window)

#### `$ bash scripts/extract_gc_content.sh ../../reference/genomes/Gencode49/GRCh38.primary_assembly.genome.fa results/intron_segments_for_phastcons.bed results/decoy_coords_for_maxent.bed > results/gc_by_decoy.tsv`

# Part 4 — Reload step-1 table, merge scores, plot, final export

```{r paths-part2}
out_dir <- "../results"
step1_tsv <- file.path(out_dir, "decoy_intron_overlap_step1.tsv")
ph100_tsv <- file.path(out_dir, "phastcons100_by_decoy.tsv")
ph470_tsv <- file.path(out_dir, "phastcons470_by_decoy.tsv")
maxent_decoy_bed <- file.path(out_dir, "maxent_decoy.bed")
maxent_canonical_bed <- file.path(out_dir, "maxent_canonical.bed")
gc_tsv <- file.path(out_dir, "gc_by_decoy.tsv")
final_tsv <- file.path(out_dir, "decoy_intron_features_final.tsv")
class_summary_tsv <- file.path(out_dir, "decoy_vs_cryptic_class_summary.tsv")
```

```{r reload-step1}
if (!file.exists(step1_tsv)) {
  stop("Missing step-1 table: ", step1_tsv, " — run Part 3 first.")
}
merged <- fread(step1_tsv, sep = "\t", header = TRUE)
cat("Reloaded rows:", nrow(merged), " cols:", ncol(merged), "\n")
```

```{r merge-phastcons}
if (file.exists(ph100_tsv)) {
  ph100 <- fread(ph100_tsv, sep = "\t", header = FALSE, col.names = c("decoyID", "phastCons100_mean"))
  merged <- merge(merged, ph100, by = "decoyID", all.x = TRUE)
} else {
  merged[, phastCons100_mean := NA_real_]
  cat("Note: missing ", ph100_tsv, " — phastCons100 not merged.\n")
}
if (file.exists(ph470_tsv)) {
  ph470 <- fread(ph470_tsv, sep = "\t", header = FALSE, col.names = c("decoyID", "phastCons470_mean"))
  merged <- merge(merged, ph470, by = "decoyID", all.x = TRUE)
} else {
  merged[, phastCons470_mean := NA_real_]
  cat("Note: missing ", ph470_tsv, " — phastCons470 not merged.\n")
}
```

```{r merge-maxent}
if (file.exists(maxent_decoy_bed)) {
  raw <- fread(maxent_decoy_bed, sep = "\t", header = FALSE)
  sc <- unique(
    data.table(decoyID = raw[[4]], MaxEnt_decoy = as.numeric(raw[[ncol(raw)]])),
    by = "decoyID"
  )
  merged <- merge(merged, sc, by = "decoyID", all.x = TRUE)
} else {
  merged[, MaxEnt_decoy := NA_real_]
  cat("Note: missing ", maxent_decoy_bed, "\n")
}
if (file.exists(maxent_canonical_bed)) {
  raw <- fread(maxent_canonical_bed, sep = "\t", header = FALSE)
  sc <- unique(
    data.table(decoyID = raw[[4]], MaxEnt_canonical = as.numeric(raw[[ncol(raw)]])),
    by = "decoyID"
  )
  merged <- merge(merged, sc, by = "decoyID", all.x = TRUE)
} else {
  merged[, MaxEnt_canonical := NA_real_]
  cat("Note: missing ", maxent_canonical_bed, "\n")
}
```

```{r merge-gc}
if (file.exists(gc_tsv)) {
  gc <- fread(gc_tsv, sep = "\t", header = TRUE)
  merged <- merge(merged, gc, by = "decoyID", all.x = TRUE)
} else {
  merged[, `:=`(intron_gc = NA_real_, site_gc_49nt = NA_real_)]
  cat("Note: missing ", gc_tsv, " — GC content not merged.\n")
}
```

## PSI count (columns with value >= 10)

```{r psi-count-v3}
meta_cols <- c(
  "GENE", "EVENT", "COORD", "LENGTH", "FullCO", "COMPLEX",
  "chr", "intron_start", "intron_end", "intron_length",
  # Numeric but not PSI; left in, they would be counted as tissues.
  "site_class", "rbpnet", "prpf8", "smb", "n_categories"
)
feat <- c(
  "decoy_coord_1based", "decoy_distance_from_canonical_5ss",
  "start", "end", "decoyID", "SpliceAI_score", "strand",
  "MaxEnt_decoy", "MaxEnt_canonical", "phastCons470_mean", "phastCons100_mean",
  "intron_gc", "site_gc_49nt"
)
psi_cols <- setdiff(names(merged), c(meta_cols, feat, "chr.1"))
psi_num <- psi_cols[sapply(merged[, ..psi_cols], is.numeric)]
merged[, psi_count_ge10 := rowSums(.SD >= 10, na.rm = TRUE), .SDcols = psi_num]
```

## mESC IRFinder retention (WT replicates) + Brain tissue PSI counts

Lift hg38 Vast-DB introns to mm10 with UCSC chain file, then overlap lifted intervals with IRFinder introns (mean `IRratio` across WT_1–WT_3).

```{r mesc-irfinder-brain}
library(rtracklayer)

sample_groups <- fread("../data/vastdb_Sample_Groups.csv")
brain_samples <- sample_groups[sample_group == "Brain", Sample]
brain_cols <- intersect(brain_samples, names(merged))
if (length(brain_cols) == 0) {
  stop("No Brain sample columns from vastdb_Sample_Groups.csv matched merged table.")
}
merged[, psi_count_ge10_brain := rowSums(.SD >= 10, na.rm = TRUE), .SDcols = brain_cols]
merged[, brain_retained := psi_count_ge10_brain >= 3L]

normalize_chr <- function(x) {
  x <- as.character(x)
  ifelse(grepl("^chr", x, ignore.case = TRUE), sub("^CHR", "chr", x), paste0("chr", x))
}

ir_dir <- "../data/Decoys/mESC_rnaIRfinder"
ir_files <- list.files(ir_dir, pattern = "WT_.*IRFinder-IR-dir\\.txt$", full.names = TRUE)
if (length(ir_files) == 0) {
  merged[, `:=`(
    mESC_IRratio_mean = NA_real_,
    mESC_IR_overlap_bp = NA_integer_,
    mESC_lift_Chr = NA_character_,
    mESC_lift_Start = NA_integer_,
    mESC_lift_End = NA_integer_,
    mESC_ir_Chr = NA_character_,
    mESC_ir_Start = NA_integer_,
    mESC_ir_End = NA_integer_,
    mESC_ir_gene = NA_character_
  )]
  cat("Note: no mESC IRFinder files in ", ir_dir, " — mESC IR not merged.\n", sep = "")
} else {
  chain_gz <- "../reference/hg38ToMm10.over.chain.gz"
  chain_plain <- "../reference/hg38ToMm10.over.chain"
  chain_url <- "https://hgdownload.soe.ucsc.edu/goldenPath/hg38/liftOver/hg38ToMm10.over.chain.gz"
  if (!file.exists(chain_gz) && !file.exists(chain_plain)) {
    dir.create(dirname(chain_gz), showWarnings = FALSE, recursive = TRUE)
    cat("Downloading ", chain_url, " ...\n", sep = "")
    download.file(chain_url, chain_gz, mode = "wb", quiet = TRUE)
  }
  if (!file.exists(chain_plain)) {
    if (!file.exists(chain_gz)) {
      stop("Missing liftOver chain file: ", chain_gz)
    }
    cat("Decompressing chain file for rtracklayer ...\n")
    system2("gunzip", c("-k", chain_gz), stdout = FALSE, stderr = FALSE)
  }

  ir_all <- rbindlist(lapply(ir_files, fread))
  ir_all[, c("ir_gene", "ensembl") := tstrsplit(Name, "/", fixed = TRUE, keep = 1:2)]
  ir_mean <- ir_all[
    , .(IRratio_mean = mean(IRratio, na.rm = TRUE), n_reps = .N),
    by = .(Chr, Start, End, ir_gene)
  ]
  ir_gr <- GRanges(
    seqnames = normalize_chr(ir_mean$Chr),
    ranges = IRanges(start = ir_mean$Start + 1L, end = ir_mean$End),
    IRratio_mean = ir_mean$IRratio_mean,
    ir_gene = ir_mean$ir_gene
  )

  chain <- import.chain(chain_plain)
  intron_dt <- unique(merged[, .(decoyID, chr, intron_start, intron_end, strand)])
  intron_gr <- GRanges(
    seqnames = normalize_chr(intron_dt$chr),
    ranges = IRanges(intron_dt$intron_start, intron_dt$intron_end),
    strand = intron_dt$strand,
    decoyID = intron_dt$decoyID
  )
  lifted_list <- liftOver(intron_gr, chain)
  lifted_gr <- unlist(lifted_list, use.names = FALSE)
  mcols(lifted_gr)$decoyID <- rep(mcols(intron_gr)$decoyID, elementNROWS(lifted_list))

  hits <- findOverlaps(lifted_gr, ir_gr)
  hit_dt <- data.table(
    decoyID = mcols(lifted_gr)$decoyID[queryHits(hits)],
    IRratio_mean = mcols(ir_gr)$IRratio_mean[subjectHits(hits)],
    mESC_ir_gene = mcols(ir_gr)$ir_gene[subjectHits(hits)],
    mESC_lift_Chr = as.character(seqnames(lifted_gr)[queryHits(hits)]),
    mESC_lift_Start = start(lifted_gr)[queryHits(hits)],
    mESC_lift_End = end(lifted_gr)[queryHits(hits)],
    mESC_ir_Chr = as.character(seqnames(ir_gr)[subjectHits(hits)]),
    mESC_ir_Start = start(ir_gr)[subjectHits(hits)],
    mESC_ir_End = end(ir_gr)[subjectHits(hits)],
    mESC_IR_overlap_bp = pmax(
      0L,
      pmin(end(lifted_gr)[queryHits(hits)], end(ir_gr)[subjectHits(hits)]) -
        pmax(start(lifted_gr)[queryHits(hits)], start(ir_gr)[subjectHits(hits)]) + 1L
    )
  )
  ir_best <- hit_dt[, .SD[which.max(mESC_IR_overlap_bp)], by = decoyID]
  n_introns <- length(intron_gr)
  n_lifted <- sum(elementNROWS(lifted_list) > 0L)
  n_lifted_segments <- length(lifted_gr)
  n_irfinder_overlap <- uniqueN(hit_dt$decoyID)

  merged <- merge(
    merged,
    ir_best[, .(
      decoyID,
      mESC_IRratio_mean = IRratio_mean,
      mESC_IR_overlap_bp,
      mESC_lift_Chr,
      mESC_lift_Start,
      mESC_lift_End,
      mESC_ir_Chr,
      mESC_ir_Start,
      mESC_ir_End,
      mESC_ir_gene
    )],
    by = "decoyID",
    all.x = TRUE
  )
  cat(
    "mESC liftOver hg38->mm10: introns lifted ", n_lifted, " / ", n_introns,
    " (", n_lifted_segments, " lifted segments); IRFinder overlap ", n_irfinder_overlap,
    "; IRratio merged ", sum(!is.na(merged$mESC_IRratio_mean)), " / ", nrow(merged),
    " (", length(ir_files), " WT replicates).\n",
    sep = ""
  )
}
cat("Brain tissues (sample_group == Brain): ", length(brain_cols), "\n", sep = "")
cat("Decoys with PSI >= 10 in >= 3 Brain tissues: ", sum(merged[site_class == "decoy", brain_retained], na.rm = TRUE), "\n", sep = "")
cat("Supported cryptics with PSI >= 10 in >= 3 Brain tissues: ", sum(merged[site_class == "cryptic_supported", brain_retained], na.rm = TRUE), "\n", sep = "")
```

```{r reorder-v3}
bed_cols <- c("chr", "start", "end", "decoyID", "SpliceAI_score", "strand")
pref <- c(
  "site_class",
  "decoy_coord_1based", "decoy_distance_from_canonical_5ss",
  "MaxEnt_decoy", "MaxEnt_canonical",
  "phastCons470_mean", "phastCons100_mean",
  "intron_length", "intron_gc", "site_gc_49nt",
  "psi_count_ge10", "psi_count_ge10_brain", "brain_retained",
  "mESC_IRratio_mean", "mESC_IR_overlap_bp",
  "mESC_lift_Chr", "mESC_lift_Start", "mESC_lift_End",
  "mESC_ir_Chr", "mESC_ir_Start", "mESC_ir_End", "mESC_ir_gene"
)
pref <- pref[pref %in% names(merged)]
setcolorder(merged, c(bed_cols, pref, setdiff(names(merged), c(bed_cols, pref))))
```

## Plot: phastCons vs mESC IR retention (WT IRFinder)

hg38 introns are lifted to mm10 (`hg38ToMm10.over.chain`) and overlapped with IRFinder intervals. Y-axis = mean `IRratio` across WT_1–WT_3 replicates.

```{r plot-phastcons-mesc-ir}
library(ggrepel)

plot_df_mesc <- merged[site_class == "decoy" & !is.na(mESC_IRratio_mean) & !is.na(GENE)]
brain_label_min <- max(1L, as.integer(ceiling(50 * length(brain_cols) / length(psi_num))))

plot_phastcons_y <- function(
  df,
  y_col,
  y_label,
  x_col,
  x_label,
  title,
  label_y_min = NULL,
  label_y_max = NULL
) {
  plot_data <- df[!is.na(get(x_col)) & !is.na(get(y_col))]
  label_plot <- plot_data[
    SpliceAI_score > 0.1 &
      !is.na(SpliceAI_score) &
      get(x_col) > 0.3
  ]
  if (!is.null(label_y_min)) {
    label_plot <- label_plot[get(y_col) >= label_y_min]
  }
  if (!is.null(label_y_max)) {
    label_plot <- label_plot[get(y_col) <= label_y_max]
  }
  label_plot <- label_plot[order(-get(y_col), -SpliceAI_score)][1:min(15L, .N)]
  if (nrow(label_plot) > 0) {
    label_plot[, label_color := ifelse(grepl("^HNRNP", GENE, ignore.case = TRUE), "salmon", "blue")]
  }

  ggplot(plot_data, aes(x = .data[[x_col]], y = .data[[y_col]], size = SpliceAI_score)) +
    geom_point(alpha = 0.5, color = "grey35") +
    geom_text_repel(
      data = label_plot,
      aes(label = GENE, color = label_color),
      size = 3,
      max.overlaps = Inf,
      show.legend = FALSE
    ) +
    scale_color_identity() +
    labs(
      title = title,
      x = x_label,
      y = y_label,
      size = "SpliceAI score"
    ) +
    theme_bw()
}

p_mesc_ph100 <- plot_phastcons_y(
  plot_df_mesc,
  "mESC_IRratio_mean",
  "mESC intron retention (mean IRratio, WT)",
  "phastCons100_mean",
  "phastCons 100-way (intron mean)",
  "phastCons 100-way vs mESC intron retention",
  label_y_min = 0.1
)
p_mesc_ph470 <- plot_phastcons_y(
  plot_df_mesc,
  "mESC_IRratio_mean",
  "mESC intron retention (mean IRratio, WT)",
  "phastCons470_mean",
  "phastCons 470-way (intron mean)",
  "phastCons 470-way vs mESC intron retention",
  label_y_min = 0.1
)

print(p_mesc_ph100)
print(p_mesc_ph470)
ggsave(file.path("../figures/phastCons470_vs_mesc_ir_retention.pdf"), p_mesc_ph470, width = 8, height = 6)
```

## Plot: phastCons vs Brain tissues with PSI > 10

Only decoys retained (PSI >= 10) in at least three Vast-DB samples with `sample_group == Brain`. Y-axis counts Brain tissues only.

```{r plot-phastcons-brain-psi}
plot_df_brain <- merged[
  site_class == "decoy" & brain_retained == TRUE & !is.na(psi_count_ge10_brain) & !is.na(GENE)
]

p_brain_ph100 <- plot_phastcons_y(
  plot_df_brain,
  "psi_count_ge10_brain",
  "Number of Brain tissues with PSI > 10",
  "phastCons100_mean",
  "phastCons 100-way (intron mean)",
  "phastCons 100-way vs Brain tissues with PSI > 10",
  label_y_min = brain_label_min
)
p_brain_ph470 <- plot_phastcons_y(
  plot_df_brain,
  "psi_count_ge10_brain",
  "Number of Brain tissues with PSI > 10",
  "phastCons470_mean",
  "phastCons 470-way (intron mean)",
  "phastCons 470-way vs Brain tissues with PSI > 10",
  label_y_min = brain_label_min
)

print(p_brain_ph100)
print(p_brain_ph470)
ggsave(file.path("../figures/phastCons470_vs_brain_psi.pdf"), p_brain_ph470, width = 8, height = 6)
```

## Plot: phastCons vs ALL Vast-DB tissues with PSI > 10

No tissue restriction and no `brain_retained` filter - every decoy with a
phastCons value and a gene symbol is plotted. The y-axis counts **all** Vast-DB
sample columns with PSI >= 10 (`psi_count_ge10`), so it spans the full tissue
panel rather than the Brain subset above.

```{r plot-phastcons-all-tissue-psi}
plot_df_all <- merged[site_class == "decoy" & !is.na(psi_count_ge10) & !is.na(GENE)]

# Label threshold scaled to the full panel, matching how brain_label_min is
# derived for the Brain subset (top ~50 of the panel by tissue count).
all_label_min <- max(1L, as.integer(ceiling(0.5 * max(plot_df_all$psi_count_ge10, na.rm = TRUE))))

cat("All-tissue plot: ", nrow(plot_df_all), " decoys across ",
    length(psi_num), " Vast-DB sample columns\n", sep = "")
cat("  label threshold (psi_count_ge10 >= ): ", all_label_min, "\n", sep = "")

p_all_ph100 <- plot_phastcons_y(
  plot_df_all,
  "psi_count_ge10",
  "Number of tissues with PSI > 10",
  "phastCons100_mean",
  "phastCons 100-way (intron mean)",
  "phastCons 100-way vs all Vast-DB tissues with PSI > 10",
  label_y_min = all_label_min
)
p_all_ph470 <- plot_phastcons_y(
  plot_df_all,
  "psi_count_ge10",
  "Number of tissues with PSI > 10",
  "phastCons470_mean",
  "phastCons 470-way (intron mean)",
  "phastCons 470-way vs all Vast-DB tissues with PSI > 10",
  label_y_min = all_label_min
)

print(p_all_ph100)
print(p_all_ph470)
dir.create("../figures", showWarnings = FALSE, recursive = TRUE)
ggsave(file.path("../figures/phastCons100_vs_all_tissue_psi.pdf"), p_all_ph100, width = 8, height = 6)
ggsave(file.path("../figures/phastCons470_vs_all_tissue_psi.pdf"), p_all_ph470, width = 8, height = 6)
```

## Plot: supported cryptics — phastCons vs ALL Vast-DB tissues with PSI > 10

The same two plots as above, for the supported cryptics: sites with no RBPnet, PRPF8 or SmB support at the site, whose intron's canonical 5' splice site is supported by RBPnet, PRPF8 or SmB.

```{r plot-phastcons-all-tissue-psi-cryptic}
plot_df_cryp <- merged[site_class == "cryptic_supported" & !is.na(psi_count_ge10) & !is.na(GENE)]
if (nrow(plot_df_cryp) == 0) stop("No supported cryptics with a PSI count and gene symbol to plot.")

cryp_label_min <- max(1L, as.integer(ceiling(0.5 * max(plot_df_cryp$psi_count_ge10, na.rm = TRUE))))
cat("Supported-cryptic plot: ", nrow(plot_df_cryp), " sites across ",
    length(psi_num), " Vast-DB sample columns\n", sep = "")
cat("  label threshold (psi_count_ge10 >= ): ", cryp_label_min, "\n", sep = "")

p_cryp_ph100 <- plot_phastcons_y(
  plot_df_cryp,
  "psi_count_ge10",
  "Number of tissues with PSI > 10",
  "phastCons100_mean",
  "phastCons 100-way (intron mean)",
  "Supported cryptics: phastCons 100-way vs all Vast-DB tissues with PSI > 10",
  label_y_min = cryp_label_min
)
p_cryp_ph470 <- plot_phastcons_y(
  plot_df_cryp,
  "psi_count_ge10",
  "Number of tissues with PSI > 10",
  "phastCons470_mean",
  "phastCons 470-way (intron mean)",
  "Supported cryptics: phastCons 470-way vs all Vast-DB tissues with PSI > 10",
  label_y_min = cryp_label_min
)

print(p_cryp_ph100)
print(p_cryp_ph470)
ggsave(file.path("../figures/phastCons100_vs_all_tissue_psi_cryptics_supported.pdf"), p_cryp_ph100, width = 8, height = 6)
ggsave(file.path("../figures/phastCons470_vs_all_tissue_psi_cryptics_supported.pdf"), p_cryp_ph470, width = 8, height = 6)
```

## Export: mESC liftOver plot table (all IR-matched loci)

Full plotting table for all decoys with a matched mESC `IRratio` after hg38→mm10 liftOver and IRFinder overlap. Use hg38 `phastCons*_mean` with `mESC_IRratio_mean` for mESC IR plots; mm10 coordinates are in `mESC_lift_*` and `mESC_ir_*`.

```{r export-mesc-plot-table}
candidates <- fread("../results/decoy_IR_candidates_mouse_brain.tsv")[, .(candidate_group, decoyID)]
plot_export <- merge(
  merged[site_class == "decoy" & !is.na(mESC_IRratio_mean)],
  candidates,
  by = "decoyID",
  all.x = TRUE
)
plot_cols <- c(
  "candidate_group", "GENE", "EVENT", "COORD", "intron_length", "chr", "strand",
  "decoyID", "start", "end", "decoy_coord_1based", "decoy_distance_from_canonical_5ss",
  "intron_start", "intron_end",
  "mESC_lift_Chr", "mESC_lift_Start", "mESC_lift_End",
  "mESC_ir_Chr", "mESC_ir_Start", "mESC_ir_End", "mESC_ir_gene",
  "mESC_IRratio_mean", "mESC_IR_overlap_bp",
  "psi_count_ge10", "psi_count_ge10_brain", "brain_retained",
  "SpliceAI_score", "phastCons100_mean", "phastCons470_mean",
  "MaxEnt_decoy", "MaxEnt_canonical"
)
plot_cols <- plot_cols[plot_cols %in% names(plot_export)]
mesc_plot_tsv <- file.path(out_dir, "decoy_intron_mesc_liftover_plot_table.tsv")
fwrite(plot_export[, ..plot_cols], mesc_plot_tsv, sep = "\t", quote = FALSE)
cat(
  "Wrote mESC plot table:", mesc_plot_tsv,
  " rows:", nrow(plot_export), "\n",
  sep = ""
)
```

## Export: IR candidate coordinates (12 highlighted genes)

```{r export-ir-candidates}
candidates <- fread("../results/decoy_IR_candidates_mouse_brain.tsv")[, .(candidate_group, decoyID)]
candidate_ids <- candidates[, decoyID]
export_dt <- merged[decoyID %in% candidate_ids]
export_dt[, brain_tissues_PSI_ge10 := apply(
  .SD, 1,
  function(v) paste(brain_cols[!is.na(v) & v >= 10], collapse = ";")
), .SDcols = brain_cols]

export_cols <- c(
  "candidate_group", "GENE", "EVENT", "COORD", "intron_length", "chr", "strand",
  "decoyID", "start", "end", "decoy_coord_1based", "decoy_distance_from_canonical_5ss",
  "intron_start", "intron_end",
  "mESC_lift_Chr", "mESC_lift_Start", "mESC_lift_End",
  "mESC_ir_Chr", "mESC_ir_Start", "mESC_ir_End", "mESC_ir_gene",
  "mESC_IRratio_mean", "mESC_IR_overlap_bp",
  "psi_count_ge10", "psi_count_ge10_brain", "brain_retained", "brain_tissues_PSI_ge10",
  "SpliceAI_score", "phastCons100_mean", "phastCons470_mean",
  "MaxEnt_decoy", "MaxEnt_canonical"
)
export_dt <- merge(candidates, export_dt, by = "decoyID", all.x = TRUE)
export_cols <- export_cols[export_cols %in% names(export_dt)]
candidate_export_tsv <- file.path(out_dir, "decoy_IRfinder_mESCliftover.tsv")
fwrite(export_dt[, ..export_cols], candidate_export_tsv, sep = "\t", quote = FALSE)
cat("Wrote candidate export:", candidate_export_tsv, " rows:", nrow(export_dt), "\n")
```

## Decoys vs supported cryptics: per-class summary and comparison figure

Every feature is computed the same way for both classes, so they can be compared directly. The summary gives n, median and IQR per class; the figure shows the distributions side by side.

```{r class-comparison}
compare_feats <- c(
  "psi_count_ge10", "psi_count_ge10_brain", "intron_length", "intron_gc", "site_gc_49nt",
  "SpliceAI_score", "decoy_distance_from_canonical_5ss",
  "MaxEnt_decoy", "MaxEnt_canonical", "phastCons100_mean", "phastCons470_mean"
)
compare_feats <- compare_feats[compare_feats %in% names(merged)]

class_summary <- merged[, .(
  n = .N,
  brain_retained_n = sum(brain_retained, na.rm = TRUE)
), by = site_class]
for (f in compare_feats) {
  q <- merged[, .(
    med = median(get(f), na.rm = TRUE),
    q1 = quantile(get(f), 0.25, na.rm = TRUE, names = FALSE),
    q3 = quantile(get(f), 0.75, na.rm = TRUE, names = FALSE),
    n_na = sum(is.na(get(f)))
  ), by = site_class]
  setnames(q, c("med", "q1", "q3", "n_na"), paste0(f, c("_median", "_q1", "_q3", "_na")))
  class_summary <- merge(class_summary, q, by = "site_class")
}
fwrite(class_summary, class_summary_tsv, sep = "\t", quote = FALSE)
cat("Wrote per-class summary:", class_summary_tsv, "\n")
print(class_summary[, c("site_class", "n", grep("_median$", names(class_summary), value = TRUE)), with = FALSE])

plot_feats <- c(
  psi_count_ge10 = "Tissues with PSI > 10",
  intron_length = "Intron length (nt, log10)",
  intron_gc = "Intron GC fraction",
  site_gc_49nt = "Site GC fraction (49 nt)",
  SpliceAI_score = "Local SpliceAI donor score",
  MaxEnt_decoy = "MaxEnt (site)",
  MaxEnt_canonical = "MaxEnt (canonical 5'SS)",
  phastCons100_mean = "phastCons 100-way (intron)",
  phastCons470_mean = "phastCons 470-way (intron)"
)
plot_feats <- plot_feats[names(plot_feats) %in% names(merged)]
long <- melt(
  cbind(merged[, .(decoyID, site_class)],
        merged[, lapply(.SD, as.numeric), .SDcols = names(plot_feats)]),
  id.vars = c("decoyID", "site_class"), variable.name = "feature", value.name = "value"
)
long <- long[!is.na(value)]
long[feature == "intron_length", value := log10(value)]
long[, feature := factor(plot_feats[as.character(feature)], levels = plot_feats)]
long[, site_class := factor(site_class, levels = c("decoy", "cryptic_supported"),
                            labels = c("Decoys", "Supported cryptics"))]

p_class <- ggplot(long, aes(x = site_class, y = value, fill = site_class)) +
  geom_violin(colour = NA, alpha = 0.4, scale = "width") +
  geom_boxplot(width = 0.18, outlier.size = 0.3, fill = "white") +
  facet_wrap(~ feature, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c("Decoys" = "#B2182B", "Supported cryptics" = "#2166AC")) +
  labs(x = NULL, y = NULL, title = "Decoys vs supported cryptics") +
  theme_bw() +
  theme(legend.position = "none", strip.text = element_text(size = 8))
print(p_class)
dir.create("../figures", showWarnings = FALSE, recursive = TRUE)
ggsave("../figures/decoy_vs_cryptic_class_comparison.pdf", p_class, width = 9, height = 9)
```

```{r write-final-v3}
fwrite(merged, final_tsv, sep = "\t", quote = FALSE)
cat("Final table:", final_tsv, " rows:", nrow(merged), " (",
    paste(merged[, .N, by = site_class][, paste0(site_class, "=", N)], collapse = ", "), ")\n")
```
