---
title: "Compile Decoy Intron Data (overlap first, then bash features)"
output: html_document
---

<!--
  Two-part workflow:
  Part 1: Load Decoys_proteincoding_splicescores.bed, overlap with Vast-DB PSI Table intron coordinates using GenomicRanges, extract strand-aware distance of decoy from canonical 5' splice site, write step1 TSV + BEDs for bash processing.
  After Part 1: Run extract_phastcons_scores.sh to calculate average phastCons scores over intron segments harboring each decoy. Use run_maxentscan_canonical.sh and run_maxentscan_decoy.sh to calculate MaxEntScan score for both the decoy and its canonical 5' splice site in bash; save TSV/BED outputs.
  Part 2: Reload step1 table, merge scores, plot, write final TSV feature table of decoys and their respective introns.

  Input splicescores: original 7-column BED (column 7 dropped on load).
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-libraries}
library(data.table)
library(GenomicRanges)
library(ggplot2)
```

# Part 1 — Load splicescores, overlap introns, write intermediate table

```{r paths-v3}
splice_bed <- "../data/Decoys/Decoys_proteincoding_splicescores.bed"
psi_table <- "../reference/PSI_TABLE-hg38.tab.gz"
out_dir <- "../results"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

step1_tsv <- file.path(out_dir, "decoy_intron_overlap_step1.tsv")
```

## Load original splice scores BED

```{r load-splice}
bed <- fread(
  splicescores_bed,
  sep = "\t",
  header = FALSE,
  col.names = c("chr", "start", "end", "decoyID", "SpliceAI_score", "strand", "counts")
)
bed[, counts := NULL]
bed[, decoyID := paste0(decoyID, "_", start)]

cat(
  "Rows:", nrow(bed), " Unique decoyID:", uniqueN(bed$decoyID),
  " Duplicates:", sum(duplicated(bed$decoyID)), "\n"
)
```

## PSI table: HsaIN only, drop `-Q` columns, parse intron coordinates

```{r load-psi-v3}
psi <- fread(psi_table, sep = "\t", header = TRUE)
psi <- psi[grepl("^HsaIN", EVENT)]
q_cols <- grep("-Q$", names(psi), value = TRUE)
if (length(q_cols) > 0) psi[, (q_cols) := NULL]

coord_parts <- tstrsplit(psi$COORD, ":", fixed = TRUE)
range_parts <- tstrsplit(coord_parts[[2]], "-", fixed = TRUE)
psi[, chr := coord_parts[[1]]]
psi[, intron_start := as.integer(range_parts[[1]])]
psi[, intron_end := as.integer(range_parts[[2]])]
psi <- psi[!is.na(chr) & !is.na(intron_start) & !is.na(intron_end)]
cat("PSI rows after filters:", nrow(psi), "\n")
```

## Overlap decoys with introns (shortest intron per decoy)

```{r overlap-v3}
decoy_gr <- GRanges(
  seqnames = bed$chr,
  ranges = IRanges(start = bed$start + 1, end = bed$end),
  strand = bed$strand
)
psi_gr <- GRanges(
  seqnames = psi$chr,
  ranges = IRanges(start = psi$intron_start, end = psi$intron_end)
)
hits <- findOverlaps(decoy_gr, psi_gr, ignore.strand = TRUE)
cat("Overlap pairs:", length(hits), "\n")

bed_hit <- bed[queryHits(hits)]
psi_hit <- psi[subjectHits(hits)]
overlap_cols <- intersect(names(bed_hit), names(psi_hit))
if (length(overlap_cols) > 0) setnames(psi_hit, overlap_cols, paste0("psi_", overlap_cols))
merged <- cbind(bed_hit, psi_hit)

merged[, intron_length := intron_end - intron_start + 1L]
setorder(merged, decoyID, intron_length)
merged <- merged[!duplicated(decoyID)]
cat("Rows after shortest-intron selection:", nrow(merged), "\n")
```

## Decoys dropped at overlap (no matching PSI intron)
Rows in `bed` with no overlap are dropped here. This block lists those decoys and writes a small BED of **all** splice scores intervals for the **genes** they belong to, for inspection (e.g. IGV).

Gene symbol is parsed from `decoyID` as `GENE_start` by stripping the trailing `_<digits>`.

```{r overlap-dropouts-genes-bed}
# Decoys present in splicescores but absent after overlap + shortest-intron dedup
dropped_ids <- setdiff(bed$decoyID, merged$decoyID)
cat(
  "Splice Scores rows:", nrow(bed), " — kept after overlap:", nrow(merged),
  " — dropped:", length(dropped_ids), "\n"
)
if (length(dropped_ids) > 0) {
  cat("Dropped decoyIDs:\n")
  print(dropped_ids)
}

bed[, gene := sub("_[0-9]+$", "", decoyID)]
genes_dropped <- unique(bed[decoyID %in% dropped_ids, gene])

cat("Genes among dropped decoys (", length(genes_dropped), "):\n", sep = "")
print(sort(genes_dropped))

dropout_bed <- file.path(out_dir, "dropped_overlap_genes_all_splicescores_decoys.bed")
bed_invest <- bed[gene %in% genes_dropped][order(gene, chr, start)]
fwrite(
  bed_invest[, .(chr, start, end, decoyID, SpliceAI_score, strand)],
  dropout_bed,
  sep = "\t",
  col.names = FALSE,
  quote = FALSE
)
cat(
  "Wrote BED (all splice scores decoys in those genes): ", dropout_bed,
  " rows:", nrow(bed_invest), "\n",
  sep = ""
)
```

## Strand-aware distance from canonical 5' splice site

```{r distance-v3}
merged[, decoy_coord_1based := start + 1L]
merged[, decoy_distance_from_canonical_5ss := ifelse(
  strand == "+",
  decoy_coord_1based - intron_start, # decoy distance caluclated from start coord
  intron_end - decoy_coord_1based # decoy distance calculated from end coord
)]
```

## Write Part-1 output (main table for bash + Part 2)

```{r write-step1}
fwrite(merged, step1_tsv, sep = "\t", quote = FALSE)
cat("Wrote:", step1_tsv, " rows:", nrow(merged), " cols:", ncol(merged), "\n")
```

## BED files for bash (phastCons + MaxEnt)

`decoy_intron_overlap_step1.tsv` alone is not enough for external tools: **phastCons** and **MaxEntScan** run in bash on **BED intervals** (and the genome FASTA for MaxEnt). This chunk writes the three BEDs that match the overlap table—**run it whenever you knit Part 1** so the commands in the “Between” section have the expected inputs. (You only skip it if you build equivalent BEDs yourself.)

- **Intron BED** — full intron span → `extract_phastcons_scores.sh` / `bigWigAverageOverBed` (mean conservation over the intron).
- **Decoy BED** — splice scores decoy interval → `run_maxentscan_decoys.sh`.
- **Canonical BED** — strand-aware canonical 5′SS (`+`: intron start; `−`: intron end) → `run_maxentscan_canonical.sh`.

```{r write-bash-input-beds}
intron_bed <- file.path(out_dir, "intron_segments_for_phastcons.bed")
decoy_bed <- file.path(out_dir, "decoy_coords_for_maxent.bed")
canonical_bed <- file.path(out_dir, "canonical_coords_for_maxent.bed")

fwrite(
  merged[, .(chr, intron_start - 1L, intron_end, decoyID, SpliceAI_score, strand)],
  intron_bed,
  sep = "\t",
  col.names = FALSE,
  eol = "\n"
)
fwrite(
  merged[, .(chr, start, end, decoyID, SpliceAI_score, strand)],
  decoy_bed,
  sep = "\t",
  col.names = FALSE,
  eol = "\n"
)
canonical_dt <- merged[, .(
  chr = chr,
  start = ifelse(strand == "+", intron_start - 1L, intron_end - 1L),
  end = ifelse(strand == "+", intron_start, intron_end),
  decoyID = decoyID,
  SpliceAI_score = SpliceAI_score,
  strand = strand
)]
fwrite(canonical_dt, canonical_bed, sep = "\t", col.names = FALSE, eol = "\n")

# Remove any CR characters so bash/awk strand parsing always sees "+" / "-" exactly for MaxEntScan
normalize_lf <- function(path) {
  txt <- readChar(path, nchars = file.info(path)$size, useBytes = TRUE)
  txt <- gsub("\r\n", "\n", txt, fixed = TRUE)
  txt <- gsub("\r", "\n", txt, fixed = TRUE)
  writeChar(txt, path, eos = NULL, useBytes = TRUE)
}
normalize_lf(intron_bed)
normalize_lf(decoy_bed)
normalize_lf(canonical_bed)

cat("Wrote:\n ", intron_bed, "\n ", decoy_bed, "\n ", canonical_bed, "\n")
```

---

## Between Part 1 and Part 2 — bash intermediate (phastCons + MaxEnt)

#### **Stop here after knitting Part 1** then run the following in the shell from `hnRNPH1_IR_MAF` (adjust paths if needed).

### phastCons (mean over each **intron** segment)

#### Use `scripts/extract_phastcons_scores.sh` (or direct `bigWigAverageOverBed`) with: phastcons bigwig and BED file input coordinates of intron of interest. Usage:

#### `$ bash scripts/extract_phastcons_scores.sh reference/hg38.phastCons100way.bw results/intron_segments_for_phastcons.bed > results/phastcons100_by_decoy.tsv`
#### `$ bash scripts/extract_phastcons_scores.sh reference/hg38.phastCons470way.bw results/intron_segments_for_phastcons.bed > results/phastcons470_by_decoy.tsv`

### MaxEntScan (decoy site vs canonical 5'SS)

#### Use `scripts/run_maxentscan_decoys.sh` and `scripts/run_maxentscan_canonical.sh` with genome FASTA as third argument. Usage:

#### `$ bash scripts/run_maxentscan_decoys.sh results/decoy_coords_for_maxent.bed results/maxent_decoy.bed ../../reference/genomes/Gencode49/GRCh38.primary_assembly.genome.fa`
#### `$ bash scripts/run_maxentscan_canonical.sh results/canonical_coords_for_maxent.bed results/maxent_canonical.bed ../../reference/genomes/Gencode49/GRCh38.primary_assembly.genome.fa`

#### Scores are appended as the **last column** of each output BED (column 4 = `decoyID`).

## Part 2 — Reload step-1 table, merge scores, plot, final export

```{r paths-part2}
out_dir <- "../results"
step1_tsv <- file.path(out_dir, "decoy_intron_overlap_step1.tsv")
ph100_tsv <- file.path(out_dir, "phastcons100_by_decoy.tsv")
ph470_tsv <- file.path(out_dir, "phastcons470_by_decoy.tsv")
maxent_decoy_bed <- file.path(out_dir, "maxent_decoy.bed")
maxent_canonical_bed <- file.path(out_dir, "maxent_canonical.bed")
final_tsv <- file.path(out_dir, "decoy_intron_features_final.tsv")
```

```{r reload-step1}
if (!file.exists(step1_tsv)) {
  stop("Missing Part-1 table: ", step1_tsv, " — run Part 1 first.")
}
merged <- fread(step1_tsv, sep = "\t", header = TRUE)
cat("Reloaded rows:", nrow(merged), " cols:", ncol(merged), "\n")
```

```{r merge-phastcons}
if (file.exists(ph100_tsv)) {
  ph100 <- fread(ph100_tsv, sep = "\t", header = FALSE, col.names = c("decoyID", "phastCons100_mean"))
  merged <- merge(merged, ph100, by = "decoyID", all.x = TRUE)
} else {
  merged[, phastCons100_mean := NA_real_]
  cat("Note: missing ", ph100_tsv, " — phastCons100 not merged.\n")
}
if (file.exists(ph470_tsv)) {
  ph470 <- fread(ph470_tsv, sep = "\t", header = FALSE, col.names = c("decoyID", "phastCons470_mean"))
  merged <- merge(merged, ph470, by = "decoyID", all.x = TRUE)
} else {
  merged[, phastCons470_mean := NA_real_]
  cat("Note: missing ", ph470_tsv, " — phastCons470 not merged.\n")
}
```

```{r merge-maxent}
if (file.exists(maxent_decoy_bed)) {
  raw <- fread(maxent_decoy_bed, sep = "\t", header = FALSE)
  sc <- unique(
    data.table(decoyID = raw[[4]], MaxEnt_decoy = as.numeric(raw[[ncol(raw)]])),
    by = "decoyID"
  )
  merged <- merge(merged, sc, by = "decoyID", all.x = TRUE)
} else {
  merged[, MaxEnt_decoy := NA_real_]
  cat("Note: missing ", maxent_decoy_bed, "\n")
}
if (file.exists(maxent_canonical_bed)) {
  raw <- fread(maxent_canonical_bed, sep = "\t", header = FALSE)
  sc <- unique(
    data.table(decoyID = raw[[4]], MaxEnt_canonical = as.numeric(raw[[ncol(raw)]])),
    by = "decoyID"
  )
  merged <- merge(merged, sc, by = "decoyID", all.x = TRUE)
} else {
  merged[, MaxEnt_canonical := NA_real_]
  cat("Note: missing ", maxent_canonical_bed, "\n")
}
```

## PSI count (columns with value >= 10)

```{r psi-count-v3}
meta_cols <- c(
  "GENE", "EVENT", "COORD", "LENGTH", "FullCO", "COMPLEX",
  "chr", "intron_start", "intron_end", "intron_length"
)
feat <- c(
  "decoy_coord_1based", "decoy_distance_from_canonical_5ss",
  "start", "end", "decoyID", "SpliceAI_score", "strand",
  "MaxEnt_decoy", "MaxEnt_canonical", "phastCons470_mean", "phastCons100_mean"
)
psi_cols <- setdiff(names(merged), c(meta_cols, feat, "chr.1"))
psi_num <- psi_cols[sapply(merged[, ..psi_cols], is.numeric)]
merged[, psi_count_ge10 := rowSums(.SD >= 10, na.rm = TRUE), .SDcols = psi_num]
```

## mESC IRFinder retention (WT replicates) + Brain tissue PSI counts

Lift hg38 Vast-DB introns to mm10 with UCSC chain file, then overlap lifted intervals with IRFinder introns (mean `IRratio` across WT_1–WT_3).

```{r mesc-irfinder-brain}
library(rtracklayer)

sample_groups <- fread("../data/vastdb_Sample_Groups.csv")
brain_samples <- sample_groups[sample_group == "Brain", Sample]
brain_cols <- intersect(brain_samples, names(merged))
if (length(brain_cols) == 0) {
  stop("No Brain sample columns from vastdb_Sample_Groups.csv matched merged table.")
}
merged[, psi_count_ge10_brain := rowSums(.SD >= 10, na.rm = TRUE), .SDcols = brain_cols]
merged[, brain_retained := psi_count_ge10_brain >= 3L]

normalize_chr <- function(x) {
  x <- as.character(x)
  ifelse(grepl("^chr", x, ignore.case = TRUE), sub("^CHR", "chr", x), paste0("chr", x))
}

ir_dir <- "../data/Decoys/mESC_rnaIRfinder"
ir_files <- list.files(ir_dir, pattern = "WT_.*IRFinder-IR-dir\\.txt$", full.names = TRUE)
if (length(ir_files) == 0) {
  merged[, `:=`(
    mESC_IRratio_mean = NA_real_,
    mESC_IR_overlap_bp = NA_integer_,
    mESC_lift_Chr = NA_character_,
    mESC_lift_Start = NA_integer_,
    mESC_lift_End = NA_integer_,
    mESC_ir_Chr = NA_character_,
    mESC_ir_Start = NA_integer_,
    mESC_ir_End = NA_integer_,
    mESC_ir_gene = NA_character_
  )]
  cat("Note: no mESC IRFinder files in ", ir_dir, " — mESC IR not merged.\n", sep = "")
} else {
  chain_gz <- "../reference/hg38ToMm10.over.chain.gz"
  chain_plain <- "../reference/hg38ToMm10.over.chain"
  chain_url <- "https://hgdownload.soe.ucsc.edu/goldenPath/hg38/liftOver/hg38ToMm10.over.chain.gz"
  if (!file.exists(chain_gz) && !file.exists(chain_plain)) {
    dir.create(dirname(chain_gz), showWarnings = FALSE, recursive = TRUE)
    cat("Downloading ", chain_url, " ...\n", sep = "")
    download.file(chain_url, chain_gz, mode = "wb", quiet = TRUE)
  }
  if (!file.exists(chain_plain)) {
    if (!file.exists(chain_gz)) {
      stop("Missing liftOver chain file: ", chain_gz)
    }
    cat("Decompressing chain file for rtracklayer ...\n")
    system2("gunzip", c("-k", chain_gz), stdout = FALSE, stderr = FALSE)
  }

  ir_all <- rbindlist(lapply(ir_files, fread))
  ir_all[, c("ir_gene", "ensembl") := tstrsplit(Name, "/", fixed = TRUE, keep = 1:2)]
  ir_mean <- ir_all[
    , .(IRratio_mean = mean(IRratio, na.rm = TRUE), n_reps = .N),
    by = .(Chr, Start, End, ir_gene)
  ]
  ir_gr <- GRanges(
    seqnames = normalize_chr(ir_mean$Chr),
    ranges = IRanges(start = ir_mean$Start + 1L, end = ir_mean$End),
    IRratio_mean = ir_mean$IRratio_mean,
    ir_gene = ir_mean$ir_gene
  )

  chain <- import.chain(chain_plain)
  intron_dt <- unique(merged[, .(decoyID, chr, intron_start, intron_end, strand)])
  intron_gr <- GRanges(
    seqnames = normalize_chr(intron_dt$chr),
    ranges = IRanges(intron_dt$intron_start, intron_dt$intron_end),
    strand = intron_dt$strand,
    decoyID = intron_dt$decoyID
  )
  lifted_list <- liftOver(intron_gr, chain)
  lifted_gr <- unlist(lifted_list, use.names = FALSE)
  mcols(lifted_gr)$decoyID <- rep(mcols(intron_gr)$decoyID, elementNROWS(lifted_list))

  hits <- findOverlaps(lifted_gr, ir_gr)
  hit_dt <- data.table(
    decoyID = mcols(lifted_gr)$decoyID[queryHits(hits)],
    IRratio_mean = mcols(ir_gr)$IRratio_mean[subjectHits(hits)],
    mESC_ir_gene = mcols(ir_gr)$ir_gene[subjectHits(hits)],
    mESC_lift_Chr = as.character(seqnames(lifted_gr)[queryHits(hits)]),
    mESC_lift_Start = start(lifted_gr)[queryHits(hits)],
    mESC_lift_End = end(lifted_gr)[queryHits(hits)],
    mESC_ir_Chr = as.character(seqnames(ir_gr)[subjectHits(hits)]),
    mESC_ir_Start = start(ir_gr)[subjectHits(hits)],
    mESC_ir_End = end(ir_gr)[subjectHits(hits)],
    mESC_IR_overlap_bp = pmax(
      0L,
      pmin(end(lifted_gr)[queryHits(hits)], end(ir_gr)[subjectHits(hits)]) -
        pmax(start(lifted_gr)[queryHits(hits)], start(ir_gr)[subjectHits(hits)]) + 1L
    )
  )
  ir_best <- hit_dt[, .SD[which.max(mESC_IR_overlap_bp)], by = decoyID]

  merged <- merge(
    merged,
    ir_best[, .(
      decoyID,
      mESC_IRratio_mean = IRratio_mean,
      mESC_IR_overlap_bp,
      mESC_lift_Chr,
      mESC_lift_Start,
      mESC_lift_End,
      mESC_ir_Chr,
      mESC_ir_Start,
      mESC_ir_End,
      mESC_ir_gene
    )],
    by = "decoyID",
    all.x = TRUE
  )
  cat(
    "mESC IRFinder (liftOver hg38->mm10): merged mean IRratio for ",
    sum(!is.na(merged$mESC_IRratio_mean)), " / ", nrow(merged),
    " decoys (", length(ir_files), " WT replicates).\n",
    sep = ""
  )
}
cat("Brain tissues (sample_group == Brain): ", length(brain_cols), "\n", sep = "")
cat("Decoys with PSI >= 10 in >= 3 Brain tissues: ", sum(merged$brain_retained, na.rm = TRUE), "\n", sep = "")
```

```{r reorder-v3}
bed_cols <- c("chr", "start", "end", "decoyID", "SpliceAI_score", "strand")
pref <- c(
  "decoy_coord_1based", "decoy_distance_from_canonical_5ss",
  "MaxEnt_decoy", "MaxEnt_canonical",
  "phastCons470_mean", "phastCons100_mean",
  "psi_count_ge10", "psi_count_ge10_brain", "brain_retained",
  "mESC_IRratio_mean", "mESC_IR_overlap_bp",
  "mESC_lift_Chr", "mESC_lift_Start", "mESC_lift_End",
  "mESC_ir_Chr", "mESC_ir_Start", "mESC_ir_End", "mESC_ir_gene"
)
pref <- pref[pref %in% names(merged)]
setcolorder(merged, c(bed_cols, pref, setdiff(names(merged), c(bed_cols, pref))))
```

## Plot: phastCons vs mESC IR retention (WT IRFinder)

hg38 introns are lifted to mm10 (`hg38ToMm10.over.chain`) and overlapped with IRFinder intervals. Y-axis = mean `IRratio` across WT_1–WT_3 replicates.

```{r plot-phastcons-mesc-ir}
library(ggrepel)

plot_df_mesc <- merged[!is.na(mESC_IRratio_mean) & !is.na(GENE)]
brain_label_min <- max(1L, as.integer(ceiling(50 * length(brain_cols) / length(psi_num))))

plot_phastcons_y <- function(
  df,
  y_col,
  y_label,
  x_col,
  x_label,
  title,
  label_y_min = NULL,
  label_y_max = NULL
) {
  plot_data <- df[!is.na(get(x_col)) & !is.na(get(y_col))]
  label_plot <- plot_data[
    SpliceAI_score > 0.1 &
      !is.na(SpliceAI_score) &
      get(x_col) > 0.3
  ]
  if (!is.null(label_y_min)) {
    label_plot <- label_plot[get(y_col) >= label_y_min]
  }
  if (!is.null(label_y_max)) {
    label_plot <- label_plot[get(y_col) <= label_y_max]
  }
  label_plot <- label_plot[order(-get(y_col), -SpliceAI_score)][1:min(15L, .N)]
  if (nrow(label_plot) > 0) {
    label_plot[, label_color := ifelse(grepl("^HNRNP", GENE, ignore.case = TRUE), "salmon", "blue")]
  }

  ggplot(plot_data, aes(x = .data[[x_col]], y = .data[[y_col]], size = SpliceAI_score)) +
    geom_point(alpha = 0.5, color = "grey35") +
    geom_text_repel(
      data = label_plot,
      aes(label = GENE, color = label_color),
      size = 3,
      max.overlaps = Inf,
      show.legend = FALSE
    ) +
    scale_color_identity() +
    labs(
      title = title,
      x = x_label,
      y = y_label,
      size = "SpliceAI score"
    ) +
    theme_bw()
}

p_mesc_ph100 <- plot_phastcons_y(
  plot_df_mesc,
  "mESC_IRratio_mean",
  "mESC intron retention (mean IRratio, WT)",
  "phastCons100_mean",
  "phastCons 100-way (intron mean)",
  "phastCons 100-way vs mESC intron retention",
  label_y_min = 0.1
)
p_mesc_ph470 <- plot_phastcons_y(
  plot_df_mesc,
  "mESC_IRratio_mean",
  "mESC intron retention (mean IRratio, WT)",
  "phastCons470_mean",
  "phastCons 470-way (intron mean)",
  "phastCons 470-way vs mESC intron retention",
  label_y_min = 0.1
)

print(p_mesc_ph100)
print(p_mesc_ph470)
```

## Plot: phastCons vs Brain tissues with PSI > 10

Only decoys retained (PSI >= 10) in at least three Vast-DB samples with `sample_group == Brain`. Y-axis counts Brain tissues only.

```{r plot-phastcons-brain-psi}
plot_df_brain <- merged[
  brain_retained == TRUE & !is.na(psi_count_ge10_brain) & !is.na(GENE)
]

p_brain_ph100 <- plot_phastcons_y(
  plot_df_brain,
  "psi_count_ge10_brain",
  "Number of Brain tissues with PSI > 10",
  "phastCons100_mean",
  "phastCons 100-way (intron mean)",
  "phastCons 100-way vs Brain tissues with PSI > 10",
  label_y_min = brain_label_min
)
p_brain_ph470 <- plot_phastcons_y(
  plot_df_brain,
  "psi_count_ge10_brain",
  "Number of Brain tissues with PSI > 10",
  "phastCons470_mean",
  "phastCons 470-way (intron mean)",
  "phastCons 470-way vs Brain tissues with PSI > 10",
  label_y_min = brain_label_min
)

print(p_brain_ph100)
print(p_brain_ph470)
```

## Export: IR candidate coordinates for reproduction

Minimal table for highlighted Mouse IR and Hs Brain IR candidates (`../results/decoy_IR_candidates_mouse_brain.tsv`). Includes hg38 decoy/intron coordinates, mm10 liftOver interval, matched IRFinder interval, mESC `IRratio`, and Brain tissues with PSI >= 10.

```{r export-ir-candidates}
candidate_ids <- fread("../results/decoy_IR_candidates_mouse_brain.tsv")[, decoyID]
export_dt <- merged[decoyID %in% candidate_ids]
export_dt[, brain_tissues_PSI_ge10 := apply(
  .SD, 1,
  function(v) paste(brain_cols[!is.na(v) & v >= 10], collapse = ";")
), .SDcols = brain_cols]

export_cols <- c(
  "candidate_group", "GENE", "EVENT", "COORD", "intron_length", "chr", "strand",
  "decoyID", "start", "end", "decoy_coord_1based", "decoy_distance_from_canonical_5ss",
  "intron_start", "intron_end",
  "mESC_lift_Chr", "mESC_lift_Start", "mESC_lift_End",
  "mESC_ir_Chr", "mESC_ir_Start", "mESC_ir_End", "mESC_ir_gene",
  "mESC_IRratio_mean", "mESC_IR_overlap_bp",
  "psi_count_ge10", "psi_count_ge10_brain", "brain_retained", "brain_tissues_PSI_ge10",
  "SpliceAI_score", "phastCons100_mean", "phastCons470_mean",
  "MaxEnt_decoy", "MaxEnt_canonical"
)
export_dt <- merge(
  fread("../results/decoy_IR_candidates_mouse_brain.tsv")[, .(candidate_group, decoyID)],
  export_dt,
  by = "decoyID",
  all.x = TRUE
)
export_cols <- export_cols[export_cols %in% names(export_dt)]
candidate_export <- export_dt[, ..export_cols]
candidate_export_tsv <- file.path(out_dir, "decoy_IR_candidates_reproduction.tsv")
fwrite(candidate_export, candidate_export_tsv, sep = "\t", quote = FALSE)
cat("Wrote candidate export:", candidate_export_tsv, " rows:", nrow(candidate_export), "\n")
```

```{r write-final-v3}
fwrite(merged, final_tsv, sep = "\t", quote = FALSE)
cat("Final table:", final_tsv, "\n")
```
