---
title: "Protein-coding filter and deep-intronic exon removal for SpliceAI splice-site inferences"
author: "Decoy Dataset Generation"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

<!--
  Two parts, with a bash step between them.

  Part 1 - STEP 1 of dataset generation.
    Keeps SpliceAI inferences (>= 0.05) that fall inside a protein-coding gene,
    on the same strand. This is the BASE dataset: everything downstream is
    derived from it. Exonic sites are deliberately retained here: step 2 splits
    them into a deep-intronic set and an exonic set (canonical + alternative
    5'SS) using Wide_canonical_splice_sites.bed, and removing exons first would
    discard the sites the exonic branch exists to capture.
    in : data/Decoys/Splice_All.filtered.05min.bed        (BED6, 794,192 sites)
    out: data/Decoys/spliceai_05min_proteincoding.bed     (BED6)

  Between: bash scripts/split_canonical_exonic.sh
    -> results/deep_intronic_splice_sites.bed, results/exonic_splice_sites.bed

  Part 2 - STEP 2b.
    Removes exon overlaps (unstranded, any transcript) from the deep-intronic
    branch only.
    in : results/deep_intronic_splice_sites.bed             (BED6)
    out: data/Decoys/deep_intronic_noexon_splice_sites.bed  (BED6)
    next: scripts/intersect_spliceai_support.sh -> decoys.bed / cryptic_sites.bed

  Knit once for Part 1 (Part 2 then stops, because its input does not exist
  yet), run the split script, and knit again.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-libraries}
library(GenomicRanges)
library(rtracklayer)
```

# Part 1 — Protein-coding filter

## Load SpliceAI inference sites

```{r load-data}
# BED6: chr, start, end, gene_name, spliceai_donor_score, strand.
# Single-nucleotide intervals (end = start + 1).
splice_sites_bed <- "../data/Decoys/Splice_All.filtered.05min.bed"
if (!file.exists(splice_sites_bed)) {
    stop(
        "Input not found: ", splice_sites_bed, "\n",
        "  Derive it from the unfiltered inference set with:\n",
        "    awk -F'\\t' '$5>=0.05' data/Decoys/Splice_All.filtered.01min.bed > ",
        splice_sites_bed
    )
}

site_bed <- read.table(splice_sites_bed,
    sep = "\t",
    header = FALSE,
    stringsAsFactors = FALSE
)

if (ncol(site_bed) != 6) {
    stop("Expected a 6-column BED, got ", ncol(site_bed), " columns.")
}
colnames(site_bed) <- c("chr", "start", "end", "name", "score", "strand")

# BED is 0-based half-open; GRanges is 1-based inclusive.
site_gr <- GRanges(
    seqnames = site_bed$chr,
    ranges = IRanges(start = site_bed$start + 1, end = site_bed$end),
    strand = site_bed$strand,
    name = site_bed$name,
    score = site_bed$score
)

cat("Loaded", length(site_gr), "SpliceAI inference sites\n")
cat("  SpliceAI score range:", min(site_bed$score), "-", max(site_bed$score), "\n")
```

## Build the protein-coding feature universe

```{r protein-coding-universe}
gtf <- import("../reference/gencode.v49.annotation.gtf.gz", format = "gtf")

# Genes annotated protein_coding, then every feature record belonging to those
# genes (gene, transcript, exon, CDS, UTR, ...). Using all features rather than
# just the gene spans keeps the behaviour identical to the previous version.
protein_coding_genes <- gtf[gtf$type == "gene" & gtf$gene_type == "protein_coding"]
cat("Protein-coding genes in annotation:", length(protein_coding_genes), "\n")

protein_coding_gene_ids <- unique(protein_coding_genes$gene_id)
protein_coding_features <- gtf[gtf$gene_id %in% protein_coding_gene_ids]
cat("Feature records belonging to those genes:", length(protein_coding_features), "\n")
```

## Keep sites overlapping a protein-coding feature (stranded)

```{r find-overlaps-protein-coding}
overlaps_protein_coding <- findOverlaps(
    query = site_gr,
    subject = protein_coding_features,
    ignore.strand = FALSE
)

cat("Overlapping pairs:", length(overlaps_protein_coding), "\n")

# Row-level selection: keep exactly the sites that themselves overlap a
# protein-coding feature.
#
# NB this is a deliberate change from the previous version, which selected on
# `name %in% unique(overlap_df$decoy_name)` - i.e. gene-symbol membership. That
# kept a site whenever ANY other site sharing its gene symbol overlapped, which
# is more permissive than intended. Both counts are printed so the difference is
# visible rather than silent.
keep_idx <- sort(unique(queryHits(overlaps_protein_coding)))
pc_bed <- site_bed[keep_idx, ]

keep_names <- unique(site_gr[keep_idx]$name)
n_by_name <- sum(site_bed$name %in% keep_names)

cat("  Sites in:                      ", nrow(site_bed), "\n")
cat("  Kept (row-level overlap):      ", nrow(pc_bed), "\n")
cat("  Would keep (gene-name, old):   ", n_by_name, "\n")
cat("  Difference attributable to key:", n_by_name - nrow(pc_bed), "\n")
cat("  Removed:                       ", nrow(site_bed) - nrow(pc_bed), "\n")
```

```{r feature-type-breakdown}
# What kind of protein-coding feature each retained site sits in. Exonic and
# intronic sites are BOTH retained here - the split happens in step 2.
ft <- protein_coding_features[subjectHits(overlaps_protein_coding)]$type
cat("Feature types hit (sites may hit several):\n")
print(sort(table(ft), decreasing = TRUE))
```

```{r deduplicate-final-bed}
before_dedup <- nrow(pc_bed)
pc_bed <- pc_bed[!duplicated(pc_bed[, 1:4]), ]
cat("Removed", before_dedup - nrow(pc_bed), "duplicate rows (chr/start/end/name)\n")
cat("Final rows:", nrow(pc_bed), "\n")
```

## Write the protein-coding base dataset

```{r save-protein-coding-results}
# BED6 in, BED6 out - the support-count columns are added later, by
# scripts/intersect_spliceai_support.sh in step 3.
out_bed <- "../data/Decoys/spliceai_05min_proteincoding.bed"
dir.create(dirname(out_bed), showWarnings = FALSE, recursive = TRUE)

write.table(pc_bed,
    file = out_bed,
    sep = "\t",
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE,
    # Force LF. Knitting from Windows RStudio against the WSL filesystem
    # otherwise writes CRLF, and the trailing \r rides along on the last column
    # where it silently breaks bedtools and every downstream numeric test.
    eol = "\n"
)

cat("Wrote", out_bed, "\n")
cat("  Rows:", nrow(pc_bed), "\n")
cat("\nNext: bash scripts/split_canonical_exonic.sh, then knit again for Part 2\n")
```

---

# Part 2 — Remove exon overlaps from the deep-intronic set

The canonical split in step 2 removes only sites within ±50 nt of an annotated canonical splice site. It leaves sites inside other exonic sequence: alternative or internal exons of the same gene, or exons of an overlapping gene on either strand. This part removes those, unstranded and against every transcript, so that the deep-intronic set is intronic in every annotated transcript before the CLIP support split. The exonic branch (`results/exonic_splice_sites.bed`) is not touched.

This includes `retained_intron` transcripts, whose exons span introns. The overlap chunk reports how many sites are removed only because of them.

```{r load-deep-intronic}
deep_intronic_path <- "../results/deep_intronic_splice_sites.bed"
if (!file.exists(deep_intronic_path)) {
    stop(
        "Input not found: ", deep_intronic_path, "\n",
        "  Part 1 has written data/Decoys/spliceai_05min_proteincoding.bed.\n",
        "  Split it, then knit again:\n",
        "    bash scripts/split_canonical_exonic.sh"
    )
}

intronic_bed <- read.table(deep_intronic_path,
    sep = "\t",
    header = FALSE,
    stringsAsFactors = FALSE
)
if (ncol(intronic_bed) != 6) {
    stop("Expected a 6-column BED, got ", ncol(intronic_bed), " columns.")
}
colnames(intronic_bed) <- c("chr", "start", "end", "name", "score", "strand")

intronic_gr <- GRanges(
    seqnames = intronic_bed$chr,
    ranges = IRanges(start = intronic_bed$start + 1, end = intronic_bed$end),
    strand = intronic_bed$strand,
    name = intronic_bed$name,
    score = intronic_bed$score
)
cat("Loaded", length(intronic_gr), "deep-intronic sites\n")
```

```{r exon-universe}
# A full knit imported the annotation in Part 1; re-import only when this part
# is run on its own.
if (!exists("gtf")) {
    gtf <- import("../reference/gencode.v49.annotation.gtf.gz", format = "gtf")
}
# Every exon record, of any gene or transcript type.
exon_gr <- gtf[gtf$type == "exon"]
cat("Exon records in annotation:", length(exon_gr), "\n")
```

```{r find-exon-overlaps-unstranded}
# Unstranded: an exon on the opposite strand (antisense or overlapping gene)
# also stops a site from being deep intronic.
exon_overlaps <- findOverlaps(
    query = intronic_gr,
    subject = exon_gr,
    ignore.strand = TRUE
)
overlapped_idx <- sort(unique(queryHits(exon_overlaps)))

same_strand <- as.character(strand(intronic_gr))[queryHits(exon_overlaps)] ==
    as.character(strand(exon_gr))[subjectHits(exon_overlaps)]
n_same <- length(unique(queryHits(exon_overlaps)[same_strand]))

not_ri <- !(exon_gr$transcript_type[subjectHits(exon_overlaps)] %in% "retained_intron")
n_ri_only <- length(overlapped_idx) - length(unique(queryHits(exon_overlaps)[not_ri]))

cat("  Deep-intronic sites:                    ", length(intronic_gr), "\n")
cat("  Sites overlapping any exon:             ", length(overlapped_idx), "\n")
cat("    with a same-strand exon:              ", n_same, "\n")
cat("    with opposite-strand exons only:      ", length(overlapped_idx) - n_same, "\n")
cat("    via retained_intron transcripts only: ", n_ri_only, "\n")
```

```{r remove-exon-overlaps}
noexon_bed <- intronic_bed[setdiff(seq_len(nrow(intronic_bed)), overlapped_idx), ]
cat("  Removed:  ", nrow(intronic_bed) - nrow(noexon_bed), "\n")
cat("  Remaining:", nrow(noexon_bed), "\n")

# The HNRNPH1 intron-4 decoy is the positive control for the whole workflow.
if (!any(noexon_bed$chr == "chr5" & noexon_bed$start == 179620582)) {
    stop("chr5:179620582 (HNRNPH1 intron-4 decoy) did not survive exon removal.")
}
```

```{r save-deep-intronic-noexon}
noexon_out <- "../data/Decoys/deep_intronic_noexon_splice_sites.bed"
write.table(noexon_bed,
    file = noexon_out,
    sep = "\t",
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE,
    eol = "\n"
)
cat("Wrote", noexon_out, "\n")
cat("  Rows:", nrow(noexon_bed), "\n")
cat(
    "\nNext (from hnRNPH1_IR_MAF):\n",
    "  bash scripts/intersect_spliceai_support.sh -w 0 -a ", sub("^\\.\\./", "", noexon_out),
    " -o results/clip_support RBPNET=... PRPF8=... SmB=...\n",
    "  mv results/clip_support_w0.bed results/decoys.bed\n",
    "  mv results/clip_support_w0_nohits.bed results/cryptic_sites.bed\n",
    sep = ""
)
```
