---
title: "Filtered Datasets Heatmaps - max1 < 0.05 and pos2 70-79"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 16, fig.height = 12)
```

## Load Libraries

```{r libraries}
library(data.table)
library(ggplot2)
library(pheatmap)
library(viridis)
library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
```

## Load Filtered Datasets

```{r load-filtered-data}
# Load the filtered datasets
max1_lt_005 <- fread("/home/mikej10/advbfx/intronretention/hnRNPH1_IR_MAF/results/splice_out_v2.5/max1_lt_005_filtered.tsv", sep = "\t", header = TRUE)
pos2_70_79 <- fread("/home/mikej10/advbfx/intronretention/hnRNPH1_IR_MAF/results/splice_out_v2.5/pos2_70_79_filtered.tsv", sep = "\t", header = TRUE)

cat("Dataset Summary:\n")
cat("  max1_lt_005: ", nrow(max1_lt_005), " rows\n")
cat("  pos2_70_79: ", nrow(pos2_70_79), " rows\n")
cat("  Unique species in max1_lt_005: ", length(unique(max1_lt_005$species)), "\n")
cat("  Unique species in pos2_70_79: ", length(unique(pos2_70_79$species)), "\n")
cat("  Unique decoyIDs in max1_lt_005: ", length(unique(max1_lt_005$decoyID)), "\n")
cat("  Unique decoyIDs in pos2_70_79: ", length(unique(pos2_70_79$decoyID)), "\n")
```

## Species Overlap Analysis

```{r species-overlap}
# Count occurrences per species in each dataset
species_max1 <- max1_lt_005 %>%
  count(species, name = "max1_count") %>%
  arrange(desc(max1_count))

species_pos2 <- pos2_70_79 %>%
  count(species, name = "pos2_count") %>%
  arrange(desc(pos2_count))

# Find species present in both datasets
species_both <- inner_join(species_max1, species_pos2, by = "species") %>%
  arrange(desc(max1_count + pos2_count))

cat("\nSpecies found in BOTH datasets:\n")
cat("  Total species in both: ", nrow(species_both), "\n")
cat("\nTop 20 species by total occurrences:\n")
print(head(species_both, 20))

# Create full comparison table
all_species <- full_join(species_max1, species_pos2, by = "species") %>%
  mutate(
    max1_count = ifelse(is.na(max1_count), 0, max1_count),
    pos2_count = ifelse(is.na(pos2_count), 0, pos2_count),
    in_both = !is.na(max1_count) & !is.na(pos2_count),
    total_count = max1_count + pos2_count
  ) %>%
  arrange(desc(total_count))

cat("\nComplete species comparison table:\n")
print(all_species)
```

## Load Species Order and Mapping

```{r load-species-order}
# Load species order from spliceai_heatmap.Rmd approach
species_order_file <- fread("/home/mikej10/advbfx/intronretention/hnRNPH1_IR_MAF/data/Newickspeciesorder.txt", header = FALSE, col.names = "tree_name")
species_mapping <- fread("/home/mikej10/advbfx/intronretention/hnRNPH1_IR_MAF/results/tree_to_clade_mapping.tsv", sep = "\t", header = TRUE)
ordered_species <- merge(species_order_file, species_mapping, by = "tree_name", all.x = TRUE, sort = FALSE)
ordered_species_ids <- ordered_species$species_id[!is.na(ordered_species$species_id)]

# Map species IDs to common names
species_id_to_common <- setNames(species_mapping$matched_common_name, species_mapping$species_id)
```

## Heatmap 1: max1 < 0.05 Dataset (showing max1 values)

```{r heatmap-max1-lt-005}
# Prepare matrix for heatmap: decoy IDs as columns, species as rows
# Extract max1 values for each decoy-species pair
max1_matrix_data <- max1_lt_005 %>%
  select(decoyID, species, max1) %>%
  pivot_wider(names_from = decoyID, values_from = max1)

# Set species as row names and convert to matrix
max1_matrix <- as.matrix(max1_matrix_data[, -1]) # Exclude species column
rownames(max1_matrix) <- max1_matrix_data$species

# Replace NA with -1 for visualization (missing data)
max1_matrix[is.na(max1_matrix)] <- -1

# Reorder rows (species) to match Newick order
available_species_max1 <- rownames(max1_matrix)
ordered_available_max1 <- intersect(ordered_species_ids, available_species_max1)
other_species_max1 <- setdiff(available_species_max1, ordered_available_max1)
final_species_order_max1 <- c(ordered_available_max1, other_species_max1)
max1_matrix <- max1_matrix[final_species_order_max1, , drop = FALSE]

# Map species IDs to common names for row labels
current_rownames_max1 <- rownames(max1_matrix)
common_names_max1 <- ifelse(current_rownames_max1 %in% names(species_id_to_common),
  species_id_to_common[current_rownames_max1],
  current_rownames_max1
)

# Create annotation for species present in both datasets (before renaming rows)
# Use original species IDs for matching
annotation_row_max1 <- data.frame(
  InBoth = ifelse(current_rownames_max1 %in% species_both$species, "Yes", "No"),
  row.names = current_rownames_max1
)

# Create annotation colors
ann_colors_max1 <- list(
  InBoth = c("Yes" = "red", "No" = "lightgrey")
)

# Rename rows to common names AFTER creating annotation
rownames(max1_matrix) <- common_names_max1
rownames(annotation_row_max1) <- common_names_max1

cat("\nMax1 < 0.05 Heatmap Matrix:\n")
cat("  Dimensions:", dim(max1_matrix), "\n")
cat("  Species (rows):", nrow(max1_matrix), "\n")
cat("  Decoy IDs (columns):", ncol(max1_matrix), "\n")
cat("  Value range:", min(max1_matrix), "to", max(max1_matrix), "\n")
cat("  -1 values (missing):", sum(max1_matrix == -1), "\n")

# Create color palette: grey for -1, viridis for 0-0.05
n_colors_max1 <- 100
color_gradient_max1 <- c("grey70", viridis(n_colors_max1))

# Create heatmap
pheatmap(
  max1_matrix,
  color = color_gradient_max1,
  breaks = c(-1.5, seq(-0.5, 0.05, length.out = n_colors_max1 + 1)),
  cluster_rows = FALSE, # Keep species in Newick order
  cluster_cols = TRUE,
  show_rownames = TRUE,
  show_colnames = FALSE, # Too many decoy IDs to show
  annotation_row = annotation_row_max1,
  annotation_colors = ann_colors_max1,
  main = "Heatmap of max1 values (max1 < 0.05 dataset) - Species Ordered by Newick Tree",
  fontsize_row = 8,
  fontsize_col = 6,
  border_color = NA,
  na_col = "grey70",
  legend_breaks = c(-1, 0, 0.01, 0.02, 0.03, 0.04, 0.05),
  legend_labels = c("Missing", "0.0", "0.01", "0.02", "0.03", "0.04", "0.05")
)
```

## Heatmap 2: pos2 70-79 Dataset (showing max2 values)

```{r heatmap-pos2-70-79}
# Prepare matrix for heatmap: decoy IDs as columns, species as rows
# Extract max2 values for each decoy-species pair
max2_matrix_data <- pos2_70_79 %>%
  select(decoyID, species, max2) %>%
  pivot_wider(names_from = decoyID, values_from = max2)

# Set species as row names and convert to matrix
max2_matrix <- as.matrix(max2_matrix_data[, -1]) # Exclude species column
rownames(max2_matrix) <- max2_matrix_data$species

# Replace NA with -1 for visualization (missing data)
max2_matrix[is.na(max2_matrix)] <- -1

# Reorder rows (species) to match Newick order
available_species_max2 <- rownames(max2_matrix)
ordered_available_max2 <- intersect(ordered_species_ids, available_species_max2)
other_species_max2 <- setdiff(available_species_max2, ordered_available_max2)
final_species_order_max2 <- c(ordered_available_max2, other_species_max2)
max2_matrix <- max2_matrix[final_species_order_max2, , drop = FALSE]

# Map species IDs to common names for row labels
current_rownames_max2 <- rownames(max2_matrix)
common_names_max2 <- ifelse(current_rownames_max2 %in% names(species_id_to_common),
  species_id_to_common[current_rownames_max2],
  current_rownames_max2
)

# Create annotation for species present in both datasets (before renaming rows)
# Use original species IDs for matching
annotation_row_max2 <- data.frame(
  InBoth = ifelse(current_rownames_max2 %in% species_both$species, "Yes", "No"),
  row.names = current_rownames_max2
)

# Create annotation colors
ann_colors_max2 <- list(
  InBoth = c("Yes" = "red", "No" = "lightgrey")
)

# Rename rows to common names AFTER creating annotation
rownames(max2_matrix) <- common_names_max2
rownames(annotation_row_max2) <- common_names_max2

cat("\nPos2 70-79 Heatmap Matrix (showing max2 values):\n")
cat("  Dimensions:", dim(max2_matrix), "\n")
cat("  Species (rows):", nrow(max2_matrix), "\n")
cat("  Decoy IDs (columns):", ncol(max2_matrix), "\n")
cat("  Value range:", min(max2_matrix), "to", max(max2_matrix), "\n")
cat("  -1 values (missing):", sum(max2_matrix == -1), "\n")

# Create color palette: grey for -1, viridis for 0-1
n_colors_max2 <- 100
color_gradient_max2 <- c("grey70", viridis(n_colors_max2))

# Create heatmap
pheatmap(
  max2_matrix,
  color = color_gradient_max2,
  breaks = c(-1.5, seq(-0.5, 1, length.out = n_colors_max2 + 1)),
  cluster_rows = FALSE, # Keep species in Newick order
  cluster_cols = TRUE,
  show_rownames = TRUE,
  show_colnames = FALSE, # Too many decoy IDs to show
  annotation_row = annotation_row_max2,
  annotation_colors = ann_colors_max2,
  main = "Heatmap of max2 values (pos2 70-79 dataset) - Species Ordered by Newick Tree",
  fontsize_row = 8,
  fontsize_col = 6,
  border_color = NA,
  na_col = "grey70",
  legend_breaks = c(-1, 0, 0.25, 0.5, 0.75, 1),
  legend_labels = c("Missing", "0.0", "0.25", "0.5", "0.75", "1.0")
)
```

## Histogram: max2 Values in pos2 70-79 Dataset

```{r histogram-max2-pos2-70-79}
# Create histogram of max2 values for pos2_70_79 dataset
p_hist_max2 <- ggplot(pos2_70_79, aes(x = max2)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7, color = "black") +
  labs(
    title = "Distribution of max2 Values (pos2 70-79 Dataset)",
    subtitle = paste("Total data points:", nrow(pos2_70_79)),
    x = "max2 (SpliceAI score)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  ) +
  # Add vertical lines for common thresholds
  geom_vline(xintercept = 0.1, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "orange", linewidth = 1) +
  geom_vline(xintercept = 0.7, linetype = "dashed", color = "blue", linewidth = 1) +
  annotate("text",
    x = 0.1, y = Inf, label = "0.1",
    vjust = 1.5, hjust = -0.1, color = "red", fontface = "bold"
  ) +
  annotate("text",
    x = 0.5, y = Inf, label = "0.5",
    vjust = 1.5, hjust = -0.1, color = "orange", fontface = "bold"
  ) +
  annotate("text",
    x = 0.7, y = Inf, label = "0.7",
    vjust = 1.5, hjust = -0.1, color = "blue", fontface = "bold"
  )

print(p_hist_max2)

# Summary statistics
cat("\nMax2 Value Summary Statistics (pos2 70-79 dataset):\n")
cat("  Mean:", sprintf("%.4f", mean(pos2_70_79$max2, na.rm = TRUE)), "\n")
cat("  Median:", sprintf("%.4f", median(pos2_70_79$max2, na.rm = TRUE)), "\n")
cat("  SD:", sprintf("%.4f", sd(pos2_70_79$max2, na.rm = TRUE)), "\n")
cat("  Min:", sprintf("%.4f", min(pos2_70_79$max2, na.rm = TRUE)), "\n")
cat("  Max:", sprintf("%.4f", max(pos2_70_79$max2, na.rm = TRUE)), "\n")
cat(
  "  Values > 0.1:", sum(pos2_70_79$max2 > 0.1, na.rm = TRUE),
  sprintf("(%.2f%%)\n", 100 * sum(pos2_70_79$max2 > 0.1, na.rm = TRUE) / nrow(pos2_70_79))
)
cat(
  "  Values > 0.5:", sum(pos2_70_79$max2 > 0.5, na.rm = TRUE),
  sprintf("(%.2f%%)\n", 100 * sum(pos2_70_79$max2 > 0.5, na.rm = TRUE) / nrow(pos2_70_79))
)
cat(
  "  Values > 0.7:", sum(pos2_70_79$max2 > 0.7, na.rm = TRUE),
  sprintf("(%.2f%%)\n", 100 * sum(pos2_70_79$max2 > 0.7, na.rm = TRUE) / nrow(pos2_70_79))
)
cat(
  "  Values > 0.8:", sum(pos2_70_79$max2 > 0.8, na.rm = TRUE),
  sprintf("(%.2f%%)\n", 100 * sum(pos2_70_79$max2 > 0.8, na.rm = TRUE) / nrow(pos2_70_79))
)
```

## Summary: Species Found in Both Datasets

```{r species-both-summary}
cat("\n")
cat(paste0(rep("=", 80), collapse = ""), "\n")
cat("SPECIES FOUND IN BOTH DATASETS\n")
cat(paste0(rep("=", 80), collapse = ""), "\n\n")

cat("Total species in max1 < 0.05 dataset: ", nrow(species_max1), "\n")
cat("Total species in pos2 70-79 dataset: ", nrow(species_pos2), "\n")
cat("Species found in BOTH datasets: ", nrow(species_both), "\n")
cat("Species only in max1 < 0.05: ", nrow(species_max1) - nrow(species_both), "\n")
cat("Species only in pos2 70-79: ", nrow(species_pos2) - nrow(species_both), "\n")

cat("\nSpecies in both datasets (sorted by total occurrences):\n")
print(species_both)
```
