---
title: "HNRNPH1 decoy element: cross-species SpliceAI position and Recount junction usage"
author: "Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    fig.width: 12
    fig.height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 8)
```

```{r load-libraries}
library(ggplot2)
library(dplyr)
library(tidyr)
library(data.table)
library(stringr)
library(ggrepel)
```

This document makes two plots for the HNRNPH1 decoy element at
**chr5:179,620,582** (minus strand):

1. **Recount junction usage** for the three 5' splice sites in this region -
   the canonical 5'SS, the decoy element, and the annotated alternative 5'SS.
2. **Cross-species position** of the strongest downstream SpliceAI donor peak
   in the decoy's intron, from the 100-way multiz alignment.

## Load cross-species SpliceAI scores for the decoy locus

```{r load-maf-spliceai}
base_path <- normalizePath("..", winslash = "/", mustWork = TRUE)

# Produced by scripts/maf_spliceai_single_locus.py, which scores ONE locus
# across every species in the multiz alignment.
maf_tsv <- file.path(base_path, "results", "hnrnph1_179620582_maf_spliceai.tsv")
if (!file.exists(maf_tsv)) {
  stop(
    "Missing: ", maf_tsv, "\n",
    "  Generate with:\n",
    "    python3 scripts/maf_spliceai_single_locus.py \\\n",
    "      --maf reference/chr5.maf.gz \\\n",
    "      --region chr5:179620038-179620940 --strand - \\\n",
    "      --out results/hnrnph1_179620582_maf_spliceai.tsv"
  )
}

hnrnph1_all <- fread(maf_tsv, sep = "\t", header = TRUE)
cat("Species scored:", nrow(hnrnph1_all), "\n")

# Attach common name and clade
species_mapping <- fread(
  file.path(base_path, "data", "tree_to_clade_mapping.tsv"),
  sep = "\t", header = TRUE
)
id2common <- setNames(species_mapping$matched_common_name, species_mapping$species_id)
id2clade <- setNames(species_mapping$clade, species_mapping$species_id)

hnrnph1_all[, common_name := ifelse(species %in% names(id2common), id2common[species], species)]
hnrnph1_all[, clade := ifelse(species %in% names(id2clade), id2clade[species], "Unknown")]

# -1 marks a species that could not be scored (absent from the alignment block,
# all-gap, or too short); drop those rather than plotting them at -1.
hnrnph1_all[max1 == -1, max1 := NA_real_]
hnrnph1_all[max2 == -1, max2 := NA_real_]
hnrnph1_all[pos2 == -1, pos2 := NA_integer_]

cat("With a scored downstream peak (max2 & pos2):",
    nrow(hnrnph1_all[!is.na(max2) & !is.na(pos2)]), "\n")
```

## Plot 1: Recount junction usage at the three 5' splice sites

```{r recount-junction-usage}
# Three 5' splice sites in the HNRNPH1 intron 4 region (all minus strand).
# The decoy is the element under investigation; the other two are the annotated
# alternatives it would compete with.
LOCI <- data.table(
  locus = factor(
    c("Canonical 5'SS", "Decoy element", "Alternative 5'SS"),
    levels = c("Canonical 5'SS", "Decoy element", "Alternative 5'SS")
  ),
  chrom = "chr5",
  position = c(179620891L, 179620582L, 179623595L),
  strand = "-"
)
print(LOCI)

# Read counts per junction from the Recount/Snaptron query.
# Produced by scripts/query_recount_loci.py; one row per
# junction-sample observation, with a `locus` column matching LOCI$locus.
recount_tsv <- file.path(base_path, "results", "hnrnph1_3loci_recount.tsv")

if (!file.exists(recount_tsv)) {
  cat(
    "NOTE: ", recount_tsv, " not present - skipping plot 1.\n",
    "  It needs the Recount/Snaptron SQLite database, which is not in this repo.\n",
    "  Once available, generate it and re-knit.\n",
    sep = ""
  )
} else {
  rc <- fread(recount_tsv, sep = "\t", header = TRUE)
  MIN_READS <- 5
  rc_f <- rc[coverage_sum >= MIN_READS]

  cat("Junction observations: ", nrow(rc), " -> ", nrow(rc_f),
      " at >= ", MIN_READS, " reads\n", sep = "")

  summary_dt <- rc_f[, .(
    n_samples = sum(samples_count, na.rm = TRUE),
    n_junctions = .N,
    median_reads = median(coverage_sum, na.rm = TRUE)
  ), by = locus]
  print(summary_dt)

  p_counts <- ggplot(rc_f, aes(x = locus, y = coverage_sum, fill = locus)) +
    geom_violin(alpha = 0.5, colour = "grey30", scale = "width") +
    geom_boxplot(width = 0.15, outlier.shape = NA, colour = "grey20", alpha = 0.9) +
    scale_y_log10(labels = scales::comma) +
    scale_fill_manual(values = c(
      "Canonical 5'SS" = "#2a78d6",
      "Decoy element" = "#A00F0F",
      "Alternative 5'SS" = "#E68A00"
    )) +
    labs(
      title = "HNRNPH1 intron 4: Recount junction read counts at three 5' splice sites",
      subtitle = paste0("Junctions with >= ", MIN_READS,
                        " reads; y-axis log10. Decoy = chr5:179,620,582"),
      x = NULL, y = "Junction read count (coverage_sum)"
    ) +
    theme_minimal(base_size = 16) +
    theme(legend.position = "none",
          axis.line = element_line(colour = "black", linewidth = 0.5))

  p_samples <- ggplot(summary_dt, aes(x = locus, y = n_samples, fill = locus)) +
    geom_col(colour = "grey20", width = 0.65) +
    geom_text(aes(label = scales::comma(n_samples)), vjust = -0.4, size = 5) +
    scale_fill_manual(values = c(
      "Canonical 5'SS" = "#2a78d6",
      "Decoy element" = "#A00F0F",
      "Alternative 5'SS" = "#E68A00"
    )) +
    labs(
      title = "Samples using each 5' splice site as a junction",
      subtitle = paste0("Recount samples with >= ", MIN_READS, " reads"),
      x = NULL, y = "Number of samples"
    ) +
    theme_minimal(base_size = 16) +
    theme(legend.position = "none",
          axis.line = element_line(colour = "black", linewidth = 0.5))

  print(p_counts)
  print(p_samples)
  dir.create(file.path(base_path, "figures"), showWarnings = FALSE, recursive = TRUE)
  ggsave(file.path(base_path, "figures", "hnrnph1_3loci_recount_readcounts.pdf"),
         p_counts, width = 10, height = 7)
  ggsave(file.path(base_path, "figures", "hnrnph1_3loci_recount_samples.pdf"),
         p_samples, width = 10, height = 7)
}
```

## Plot 2: Cross-species position of the decoy element



```{r scatter-plot-hnrnph1-max2-pos2}
# hnrnph1_all is loaded and annotated in the load-maf-spliceai chunk above;
# species_mapping / clade lookups are already applied there.

# Persist the per-species table for this decoy (same path as before, so any
# downstream consumer keeps working).
fwrite(
  hnrnph1_all,
  file = file.path(base_path, "data", "Decoys", "hnrnph1_spliceaispecies.tsv"),
  sep = "\t",
  eol = "\n"
)

# Subset for the positional scatterplot
hnrnph1_data_max2 <- hnrnph1_all %>% filter(!is.na(max2) & !is.na(pos2))

cat("\nHNRNPH1_179620582 data:\n")
cat("  Species rows saved:", nrow(hnrnph1_all), "\n")
cat("  Species with max2 and pos2:", nrow(hnrnph1_data_max2), "\n")

# --- Collapse the source clades into the four plotted groups --------------------
# The mapping file's clade column splits mammals across five labels. For this
# figure all mammals are shown together, split only by placental status:
#   Placental (Eutheria) = Afrotheria + Euarchontoglires + Laurasiatheria +
#                          Primate, plus dasNov3 (Armadillo). NB: Armadillo sits
#                          in the catch-all "Mammal" clade but is a xenarthran,
#                          i.e. placental - it must not fall in with the pouched
#                          mammals below.
#   Other mammals        = the marsupials and the monotreme (Opossum, Tasmanian
#                          devil, Wallaby, Platypus).
# Sarcopterygii (alligator, turtles, lizard, coelacanth, X. tropicalis) is
# dropped from this figure entirely. Aves and Fish are retained as-is.
PLACENTAL_CLADES <- c("Afrotheria", "Euarchontoglires", "Laurasiatheria", "Primate")
PLACENTAL_EXTRA <- c("dasNov3") # Armadillo (Xenarthra)

hnrnph1_data_max2 <- hnrnph1_data_max2 %>%
  mutate(
    group = case_when(
      clade %in% PLACENTAL_CLADES | species %in% PLACENTAL_EXTRA ~ "Placental mammals",
      clade == "Mammal" ~ "Other mammals",
      clade == "Aves" ~ "Aves",
      clade == "Fish" ~ "Fish",
      TRUE ~ NA_character_ # Sarcopterygii -> dropped
    )
  ) %>%
  filter(!is.na(group)) %>%
  mutate(group = factor(group,
    levels = c("Placental mammals", "Other mammals", "Aves", "Fish")
  ))

cat("\nSpecies per plotted group (Sarcopterygii dropped):\n")
print(table(hnrnph1_data_max2$group))

# Only these species carry a text label; every other point is unlabelled.
LABEL_SPECIES <- c(
  "Cat", "Pig", "Human", "Rat", "Cow",
  "Gorilla", "Chimp", "Baboon", "Mouse"
)
label_data <- hnrnph1_data_max2 %>% filter(common_name %in% LABEL_SPECIES)
cat("  Labelled species found:", nrow(label_data), "of", length(LABEL_SPECIES), "\n")
missing_labels <- setdiff(LABEL_SPECIES, label_data$common_name)
if (length(missing_labels) > 0) {
  cat(
    "  WARNING - requested label absent from data:",
    paste(missing_labels, collapse = ", "), "\n"
  )
}

# Palette validated for colour-vision deficiency (OKLab dE, Machado 2009 sev 1.0,
# all pairs). Placental/other mammals are deliberately adjacent hues (red/orange)
# so the two read as one mammalian block. Two pairs sit in WARN bands - the orange
# is below 3:1 against a white surface, and red-vs-green falls to dE 6.7 under
# deuteranopia - so `group` is redundantly encoded by SHAPE as well as fill, and
# every point carries a dark stroke. Do not drop the shape scale to "tidy up".
GROUP_FILL <- c(
  "Placental mammals" = "#A00F0F", # dark red
  "Other mammals"     = "#E68A00", # dark orange
  "Aves"              = "#2a78d6",
  "Fish"              = "#008300"
)
GROUP_SHAPE <- c(
  "Placental mammals" = 21, # circle
  "Other mammals"     = 24, # triangle
  "Aves"              = 22, # square
  "Fish"              = 23  # diamond
)

if (nrow(hnrnph1_data_max2) > 0) {
  p_hnrnph1_max2_pos2 <- ggplot(
    hnrnph1_data_max2,
    aes(x = pos2, y = max2, fill = group, shape = group)
  ) +
    geom_point(size = 4, alpha = 0.9, colour = "grey20", stroke = 0.6) +
    # The nine labelled species are redrawn larger with a black ring. In the
    # dense mammalian column a leader line can otherwise pass close enough to a
    # neighbouring point to be misread as pointing at it; this makes the label
    # target unambiguous.
    # show.legend = FALSE: this layer must not touch the legend keys, or the
    # "Placental mammals" key inherits the black ring and reads as if the ring
    # meant "placental" rather than "labelled".
    geom_point(
      data = label_data, size = 6, alpha = 1, colour = "black",
      stroke = 1.2, show.legend = FALSE
    ) +
    # Labels for the nine focus species only
    geom_text_repel(
      data = label_data,
      aes(label = common_name),
      size = 6,
      fontface = "bold",
      colour = "grey10",
      min.segment.length = 0,
      segment.size = 0.4,
      segment.color = "grey40",
      box.padding = 0.9,
      point.padding = 0.5,
      max.overlaps = Inf,
      show.legend = FALSE
    ) +
    scale_fill_manual(values = GROUP_FILL, name = NULL, drop = FALSE) +
    scale_shape_manual(values = GROUP_SHAPE, name = NULL, drop = FALSE) +
    # fill and shape map to the same variable, so ggplot merges them into a
    # single legend; set override.aes on one guide only or it warns.
    guides(fill = guide_legend(override.aes = list(size = 6))) +
    labs(
      title = "HNRNPH1_179620582 Decoy Element Retains Locus Position Across Species",
      subtitle = "SpliceAI donor scores calculated across whole intron multiple alignment file for maximum position",
      x = "Loci distance from canonical 5' splice site (nt)",
      y = "Consensus Sequence Strength (SpliceAI Donor Score)"
    ) +
    theme_minimal(base_size = 18) +
    theme(
      plot.title = element_text(size = 22, face = "bold"),
      plot.subtitle = element_text(size = 15, colour = "grey30"),
      axis.title = element_text(size = 18),
      axis.text = element_text(size = 16, colour = "grey20"),
      legend.position = "right",
      legend.text = element_text(size = 16),
      legend.key.height = grid::unit(1.4, "lines"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey90", linewidth = 0.4),
      axis.line.x = element_line(colour = "black", linewidth = 0.5),
      axis.line.y = element_line(colour = "black", linewidth = 0.5)
    )

  print(p_hnrnph1_max2_pos2)
  pdf(file.path(base_path, "figures", "hnrnph1_max2_pos2_scatterplot.pdf"), width = 16, height = 11)
  print(p_hnrnph1_max2_pos2)
  dev.off()
  ggsave(
    file.path(base_path, "figures", "hnrnph1_max2_pos2_scatterplot.png"),
    p_hnrnph1_max2_pos2,
    width = 16, height = 11, dpi = 300, bg = "white"
  )
  cat("  Output PDF: ", file.path(base_path, "results", "hnrnph1_max2_pos2_scatterplot.pdf"), "\n", sep = "")

  # Print summary statistics
  cat("\nSummary statistics (max2 plot):\n")
  cat("  Mean max2:", round(mean(hnrnph1_data_max2$max2, na.rm = TRUE), 4), "\n")
  cat("  Median max2:", round(median(hnrnph1_data_max2$max2, na.rm = TRUE), 4), "\n")
  cat("  Min max2:", round(min(hnrnph1_data_max2$max2, na.rm = TRUE), 4), "\n")
  cat("  Max max2:", round(max(hnrnph1_data_max2$max2, na.rm = TRUE), 4), "\n")
  cat("  Mean pos2:", round(mean(hnrnph1_data_max2$pos2, na.rm = TRUE), 2), "\n")
  cat("  Median pos2:", round(median(hnrnph1_data_max2$pos2, na.rm = TRUE), 2), "\n")
  cat("  Min pos2:", min(hnrnph1_data_max2$pos2, na.rm = TRUE), "\n")
  cat("  Max pos2:", max(hnrnph1_data_max2$pos2, na.rm = TRUE), "\n")
} else {
  cat("  No data found for HNRNPH1_179620582 with both max2 and pos2\n")
}
```
