---
title: "Intron summary table: all Vast-DB introns classified by decoy and cryptic content"
output: html_document
params:
  fasta: "../../../reference/genomes/Gencode49/GRCh38.primary_assembly.genome.fa"
---

<!--
  Builds one row per Vast-DB HsaIN intron from PSI_TABLE-hg38.tab.gz, marks the
  introns that hold decoys and/or supported cryptics from
  results/decoy_intron_features_final.tsv, and adds intron length, tissue
  counts with PSI >= 10, and GC content of the intron relative to its flanking
  exons. Run after Part 4 of compile_decoy_intron_data.Rmd.

  out: results/intron_summary_table.tsv
       figures/intron_class_comparison.pdf
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-libraries}
library(data.table)
library(ggplot2)
```

```{r paths}
psi_table <- "../reference/PSI_TABLE-hg38.tab.gz"
features_tsv <- "../results/decoy_intron_features_final.tsv"
sample_groups_csv <- "../data/vastdb_Sample_Groups.csv"
fasta <- params$fasta
out_tsv <- "../results/intron_summary_table.tsv"
fig_dir <- "../figures"
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

if (!file.exists(fasta)) stop("Genome FASTA not found: ", fasta, " (set params$fasta)")
if (!file.exists(paste0(fasta, ".fai"))) stop("FASTA index not found: ", fasta, ".fai")
if (Sys.which("bedtools") == "") stop("bedtools not on PATH")

HNRNPH1_ID <- "HNRNPH1_179620582"
CLASS_LEVELS <- c("none", "decoy", "cryptic", "both")
CLASS_LABEL <- c(none = "All other introns", decoy = "Decoy", cryptic = "Supported cryptic", both = "Both")
CLASS_COLOR <- c(none = "grey60", decoy = "#B2182B", cryptic = "#2166AC", both = "#762A83")
```

## All Vast-DB introns

`COORD` gives the intron; `FullCO` gives the two flanking exons in transcript order (`chr:up_start-up_end=down_start-down_end:strand`), so the upstream exon is the first pair on both strands.

```{r load-introns}
psi <- fread(psi_table, sep = "\t", header = TRUE)
psi <- psi[grepl("^HsaIN", EVENT)]
q_cols <- grep("-Q$", names(psi), value = TRUE)
if (length(q_cols) > 0) psi[, (q_cols) := NULL]

coord <- tstrsplit(psi$COORD, "[:-]")
psi[, `:=`(chr = coord[[1]], intron_start = as.integer(coord[[2]]), intron_end = as.integer(coord[[3]]))]
psi[, intron_length := intron_end - intron_start + 1L]

# Strand is the last character; it must come off before splitting on "-".
psi[, strand := sub(".*:", "", FullCO)]
fc <- tstrsplit(sub(":[+-]$", "", psi$FullCO), "[:=-]")
psi[, `:=`(
  up_exon_start = as.integer(fc[[2]]), up_exon_end = as.integer(fc[[3]]),
  down_exon_start = as.integer(fc[[4]]), down_exon_end = as.integer(fc[[5]])
)]
psi <- psi[!is.na(intron_start) & !is.na(up_exon_start) & !is.na(down_exon_start) & strand %in% c("+", "-")]
psi[, up_exon_length := up_exon_end - up_exon_start + 1L]
psi[, down_exon_length := down_exon_end - down_exon_start + 1L]

meta_cols <- c("GENE", "EVENT", "COORD", "LENGTH", "FullCO", "COMPLEX", "chr", "intron_start", "intron_end",
               "intron_length", "up_exon_start", "up_exon_end", "down_exon_start", "down_exon_end", "strand",
               "up_exon_length", "down_exon_length")
psi_cols <- setdiff(names(psi), meta_cols)
psi_cols <- psi_cols[sapply(psi[, ..psi_cols], is.numeric)]
psi[, psi_count_ge10 := rowSums(.SD >= 10, na.rm = TRUE), .SDcols = psi_cols]
psi[, psi_n_measured := rowSums(!is.na(.SD)), .SDcols = psi_cols]

sample_groups <- fread(sample_groups_csv)
brain_cols <- intersect(sample_groups[sample_group == "Brain", Sample], psi_cols)
psi[, psi_count_ge10_brain := rowSums(.SD >= 10, na.rm = TRUE), .SDcols = brain_cols]

cat("Vast-DB introns:", nrow(psi), "  tissue columns:", length(psi_cols), "  brain columns:", length(brain_cols), "\n")
stopifnot(!anyDuplicated(psi$EVENT))
```

## Mark introns that hold decoys and supported cryptics

Each site in the feature table is assigned to the shortest Vast-DB intron overlapping it (`EVENT`). Site IDs are `GENE_start`.

```{r mark-sites}
feat <- fread(features_tsv, sep = "\t", header = TRUE, select = c("decoyID", "site_class", "EVENT", "start"))
stopifnot(all(feat$EVENT %in% psi$EVENT))

per_intron <- feat[, .(
  n_decoys = sum(site_class == "decoy"),
  n_cryptics = sum(site_class == "cryptic_supported"),
  decoy_ids = paste(sort(decoyID[site_class == "decoy"]), collapse = ";"),
  cryptic_ids = paste(sort(decoyID[site_class == "cryptic_supported"]), collapse = ";")
), by = EVENT]

intron <- merge(psi[, ..meta_cols][, c(.SD, list(psi_count_ge10 = psi$psi_count_ge10,
                                                   psi_count_ge10_brain = psi$psi_count_ge10_brain,
                                                   psi_n_measured = psi$psi_n_measured))],
                per_intron, by = "EVENT", all.x = TRUE)
intron[is.na(n_decoys), `:=`(n_decoys = 0L, n_cryptics = 0L, decoy_ids = "", cryptic_ids = "")]
intron[, intron_class := fifelse(n_decoys > 0 & n_cryptics > 0, "both",
                         fifelse(n_decoys > 0, "decoy",
                         fifelse(n_cryptics > 0, "cryptic", "none")))]
intron[, intron_class := factor(intron_class, levels = CLASS_LEVELS)]

class_counts <- intron[, .(n = .N), by = intron_class][order(intron_class)]
class_counts[, pct_of_all_introns := round(100 * n / nrow(intron), 2)]
cat("Introns by class (of", nrow(intron), "Vast-DB introns):\n"); print(class_counts)
with_site <- intron[intron_class != "none", .N]
cat(sprintf("\nIntrons holding at least one site: %d (%.2f%%); of these, %.1f%% decoy only, %.1f%% cryptic only, %.1f%% both\n",
  with_site, 100 * with_site / nrow(intron),
  100 * class_counts[intron_class == "decoy", n] / with_site,
  100 * class_counts[intron_class == "cryptic", n] / with_site,
  100 * class_counts[intron_class == "both", n] / with_site))
cat(sprintf("Of the %d introns with a decoy, %.1f%% also hold a supported cryptic; of the %d with a cryptic, %.1f%% also hold a decoy\n",
  intron[n_decoys > 0, .N], 100 * class_counts[intron_class == "both", n] / intron[n_decoys > 0, .N],
  intron[n_cryptics > 0, .N], 100 * class_counts[intron_class == "both", n] / intron[n_cryptics > 0, .N]))
fwrite(class_counts, "../results/intron_class_counts.tsv", sep = "\t", quote = FALSE)
cat("\nSites per intron (rows = decoys, columns = cryptics):\n")
print(table(decoys = intron$n_decoys, cryptics = intron$n_cryptics))
cat("\nSites in the feature table:", nrow(feat), " assigned to", nrow(per_intron), "introns\n")

hn_event <- feat[decoyID == HNRNPH1_ID, EVENT]
stopifnot(length(hn_event) == 1)
cat("\nHNRNPH1 intron:", hn_event, "\n")
print(intron[EVENT == hn_event, .(GENE, COORD, strand, intron_length, intron_class, n_decoys, n_cryptics, decoy_ids, cryptic_ids, psi_count_ge10)])
```

## GC content: intron relative to flanking exons

`bedtools nuc` on the intron and on each flanking exon. `exon_gc` is the length-weighted GC of the two exons together; `gc_ratio` is `intron_gc / exon_gc`.

```{r gc-content}
tmp <- tempfile("gc_")
dir.create(tmp)
write_bed <- function(dt, start1, end1, path) {
  fwrite(dt[, .(chr, get(start1) - 1L, get(end1), EVENT, 0L, strand)], path,
         sep = "\t", col.names = FALSE, quote = FALSE, eol = "\n")
}
write_bed(intron, "intron_start", "intron_end", file.path(tmp, "intron.bed"))
write_bed(intron, "up_exon_start", "up_exon_end", file.path(tmp, "up.bed"))
write_bed(intron, "down_exon_start", "down_exon_end", file.path(tmp, "down.bed"))

# bedtools nuc on BED6 appends: 7 pct_at, 8 pct_gc, 9 num_A, 10 num_C, 11 num_G, 12 num_T, 13 num_N, 14 num_oth, 15 seq_len
nuc <- function(bed) {
  out <- fread(cmd = paste("bedtools nuc -fi", shQuote(fasta), "-bed", shQuote(bed)), sep = "\t", header = TRUE)
  setnames(out, c("chr", "start", "end", "EVENT", "score", "strand", "pct_at", "pct_gc",
                  "num_A", "num_C", "num_G", "num_T", "num_N", "num_oth", "seq_len"))
  out[, .(EVENT, gc = num_C + num_G, len = seq_len - num_N)]
}
gi <- nuc(file.path(tmp, "intron.bed")); gu <- nuc(file.path(tmp, "up.bed")); gd <- nuc(file.path(tmp, "down.bed"))
stopifnot(nrow(gi) == nrow(intron), nrow(gu) == nrow(intron), nrow(gd) == nrow(intron))
gc <- Reduce(function(a, b) merge(a, b, by = "EVENT"), list(
  gi[, .(EVENT, intron_gc = gc / len)],
  gu[, .(EVENT, up_exon_gc = gc / len, up_gc = gc, up_len = len)],
  gd[, .(EVENT, down_exon_gc = gc / len, down_gc = gc, down_len = len)]
))
gc[, exon_gc := (up_gc + down_gc) / (up_len + down_len)]
gc[, gc_ratio := intron_gc / exon_gc]
intron <- merge(intron, gc[, .(EVENT, intron_gc, up_exon_gc, down_exon_gc, exon_gc, gc_ratio)], by = "EVENT")
unlink(tmp, recursive = TRUE)

cat("GC ratio (intron / flanking exons), median by class:\n")
print(intron[, .(n = .N, median_gc_ratio = median(gc_ratio, na.rm = TRUE)), by = intron_class][order(intron_class)])
```

## Write the summary table

```{r write-table}
setcolorder(intron, c("GENE", "EVENT", "COORD", "chr", "intron_start", "intron_end", "strand", "intron_length",
                      "intron_class", "n_decoys", "n_cryptics", "decoy_ids", "cryptic_ids",
                      "psi_count_ge10", "psi_count_ge10_brain", "psi_n_measured",
                      "intron_gc", "exon_gc", "gc_ratio", "up_exon_gc", "down_exon_gc",
                      "up_exon_start", "up_exon_end", "up_exon_length",
                      "down_exon_start", "down_exon_end", "down_exon_length"))
setorder(intron, chr, intron_start)
fwrite(intron, out_tsv, sep = "\t", quote = FALSE, eol = "\n")
cat("Wrote", out_tsv, " rows:", nrow(intron), " cols:", ncol(intron), "\n")
```

## Intron classes compared

Retention (tissues with PSI >= 10), intron length, GC content and GC ratio by intron class, with all other introns as the control. Per-class n, median and IQR are printed; the HNRNPH1 intron 4 is marked.

```{r class-comparison}
plot_feats <- c(
  psi_count_ge10 = "Tissues with PSI > 10",
  psi_count_ge10_brain = "Brain tissues with PSI > 10",
  intron_length = "Intron length (nt, log10)",
  intron_gc = "Intron GC fraction",
  exon_gc = "Flanking-exon GC fraction",
  gc_ratio = "GC ratio (intron / flanking exons)"
)
summary_tbl <- rbindlist(lapply(names(plot_feats), function(f) {
  intron[, .(feature = f, n = .N,
             median = as.numeric(median(get(f), na.rm = TRUE)),
             q1 = as.numeric(quantile(get(f), 0.25, na.rm = TRUE, names = FALSE)),
             q3 = as.numeric(quantile(get(f), 0.75, na.rm = TRUE, names = FALSE))), by = intron_class]
}))
print(dcast(summary_tbl, feature ~ intron_class, value.var = "median"))
fwrite(summary_tbl, "../results/intron_class_summary.tsv", sep = "\t", quote = FALSE)

long <- melt(intron[, c("EVENT", "intron_class", names(plot_feats)), with = FALSE],
             id.vars = c("EVENT", "intron_class"), variable.name = "feature", value.name = "value")
long <- long[!is.na(value)]
long[feature == "intron_length", value := log10(value)]
long[, feature := factor(plot_feats[as.character(feature)], levels = plot_feats)]
long[, class_label := factor(CLASS_LABEL[as.character(intron_class)], levels = CLASS_LABEL)]
hn <- long[EVENT == hn_event]

p_class <- ggplot(long, aes(x = class_label, y = value, fill = class_label)) +
  geom_violin(colour = NA, alpha = 0.45, scale = "width") +
  geom_boxplot(width = 0.16, outlier.size = 0.2, fill = "white") +
  geom_point(data = hn, shape = 23, size = 3.2, fill = "black", colour = "white") +
  geom_text(data = hn, label = "HNRNPH1", vjust = -1, size = 2.8, fontface = "bold") +
  facet_wrap(~ feature, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = setNames(CLASS_COLOR, CLASS_LABEL)) +
  labs(x = NULL, y = NULL, title = "Vast-DB introns by decoy / supported-cryptic content") +
  theme_bw() +
  theme(legend.position = "none", strip.text = element_text(size = 8),
        axis.text.x = element_text(angle = 25, hjust = 1, size = 8))
print(p_class)
ggsave(file.path(fig_dir, "intron_class_comparison.pdf"), p_class, width = 10, height = 7)
```

```{r retention-violin}
counts <- intron[, .N, by = intron_class][order(intron_class)]
xlab <- paste0(CLASS_LABEL[as.character(counts$intron_class)], "\n(n = ", format(counts$N, big.mark = ","), ")")
hn_row <- intron[EVENT == hn_event]
p_ret <- ggplot(intron, aes(x = intron_class, y = psi_count_ge10, fill = intron_class)) +
  geom_violin(colour = NA, alpha = 0.45, scale = "width", cut = 0) +
  geom_boxplot(width = 0.16, outlier.shape = NA, fill = "white") +
  geom_point(data = hn_row, shape = 23, size = 3.5, fill = "black", colour = "white") +
  annotate("text", x = as.integer(hn_row$intron_class), y = hn_row$psi_count_ge10, label = "HNRNPH1",
           vjust = -1, size = 3, fontface = "bold") +
  scale_x_discrete(labels = xlab) +
  scale_fill_manual(values = CLASS_COLOR) +
  labs(x = NULL, y = "Tissues with PSI > 10", title = "Intron retention by intron class") +
  theme_bw() + theme(legend.position = "none")
print(p_ret)
ggsave(file.path(fig_dir, "intron_class_retention.pdf"), p_ret, width = 7, height = 5)
```
