---
title: "SpliceAI Validation Plots - Position and Score Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 16, fig.height = 12)
```

## Load Libraries

```{r libraries}
library(data.table)
library(ggplot2)
library(pheatmap)
library(viridis)
library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(GenomicRanges)
```

## QC: Validate BED File Coordinates

```{r qc-bed-coordinates}
# Read the BED file
bed_data <- fread("/home/mikej10/advbfx/intronretention/hnRNPH1_IR_MAF/data/DecoysInConservedGenes50thpercentilev2.bed", 
                  sep = "\t", 
                  header = FALSE,
                  col.names = c("chr", "start", "end", "id", "score", "strand"))

cat("BED file loaded:\n")
cat("  Total records:", nrow(bed_data), "\n")

# Extract coordinate after underscore from column 4 (id)
bed_data$coord_from_id <- as.integer(str_extract(bed_data$id, "(?<=_)\\d+$"))

# Check for any missing coordinates
missing_coords <- sum(is.na(bed_data$coord_from_id))
if (missing_coords > 0) {
  cat("  WARNING: ", missing_coords, " records have missing coordinates in ID\n")
}

# Create GenomicRanges object for the BED intervals
bed_gr <- GRanges(
  seqnames = bed_data$chr,
  ranges = IRanges(start = bed_data$start, end = bed_data$end),
  strand = bed_data$strand,
  id = bed_data$id,
  coord_from_id = bed_data$coord_from_id
)

# Check if coordinate from ID falls within the start-end range
bed_data$coord_in_range <- bed_data$coord_from_id >= bed_data$start & 
                           bed_data$coord_from_id <= bed_data$end

# Summary statistics
total_records <- nrow(bed_data)

cat("  Total records:", total_records, "\n")
```

## Load and Parse Data

```{r load-data}
# Read the TSV file
spliceai_data <- fread("/home/mikej10/advbfx/intronretention/hnRNPH1_IR_MAF/results/splice_out_v2.5/all_spliceai_scores_v2.5.tsv", sep = "\t", header = TRUE)

cat("Dataset loaded:\n")
cat("  Number of decoy IDs:", nrow(spliceai_data), "\n")
cat("  Number of species columns:", ncol(spliceai_data) - 1, "\n")

# Function to parse [max1][pos1][max2][pos2][d] format (v2.5)
parse_score_format <- function(value) {
  # Match pattern: [number][number][number][number][number]
  match_result <- str_match(value, "\\[([\\d.-]+)\\]\\[(\\d+)\\]\\[([\\d.-]+)\\]\\[(\\d+)\\]\\[([\\d.-]+)\\]")
  if (is.na(match_result[1])) {
    return(list(max1 = NA, pos1 = NA, max2 = NA, pos2 = NA, d = NA))
  }
  return(list(
    max1 = as.numeric(match_result[2]),
    pos1 = as.integer(match_result[3]),
    max2 = as.numeric(match_result[4]),
    pos2 = as.integer(match_result[5]),
    d = as.numeric(match_result[6])
  ))
}

# Extract decoy IDs
decoy_ids <- spliceai_data$decoyID

# Get species columns (all except decoyID)
species_cols <- setdiff(names(spliceai_data), "decoyID")

cat("\nParsing score data for all species...\n")

# Parse all data into long format
parsed_data_list <- list()

for (species in species_cols) {
  parsed_scores <- lapply(spliceai_data[[species]], parse_score_format)
  
  # Convert to data.table
  species_data <- data.table(
    decoyID = decoy_ids,
    species = species,
    max1 = sapply(parsed_scores, function(x) x$max1),
    pos1 = sapply(parsed_scores, function(x) x$pos1),
    max2 = sapply(parsed_scores, function(x) x$max2),
    pos2 = sapply(parsed_scores, function(x) x$pos2),
    d = sapply(parsed_scores, function(x) x$d)
  )
  
  parsed_data_list[[species]] <- species_data
}

# Combine all species data
all_data <- rbindlist(parsed_data_list)

# Check for duplicate decoy IDs in the original dataset
duplicate_decoy_ids <- spliceai_data$decoyID[duplicated(spliceai_data$decoyID)]
if (length(duplicate_decoy_ids) > 0) {
  cat("\n⚠️  WARNING: Found duplicate decoy IDs in the dataset:\n")
  cat("  Number of duplicates:", length(duplicate_decoy_ids), "\n")
  cat("  Duplicate decoy IDs:\n")
  for (dup_id in duplicate_decoy_ids) {
    cat(sprintf("    %s\n", dup_id))
  }
  cat("\n  These should be checked in the original BED file.\n")
} else {
  cat("\n✓ No duplicate decoy IDs found in the dataset.\n")
}

# Filter out -1 values for analysis
all_data_valid <- all_data[max1 != -1.0 & pos1 != -1 & max2 != -1.0 & pos2 != -1]

cat("\nData parsing complete:\n")
cat("  Total data points:", nrow(all_data), "\n")
cat("  Valid data points (excluding -1):", nrow(all_data_valid), "\n")
cat("  Unique decoy IDs:", length(unique(all_data_valid$decoyID)), "\n")
cat("  Unique species:", length(unique(all_data_valid$species)), "\n")

# Extract and save filtered datasets
# 1. Extract rows where pos2 is between 70-79
pos2_70_79 <- all_data_valid[pos2 >= 70 & pos2 <= 79]
cat("\nExtracting rows where pos2 is between 70-79:\n")
cat("  Number of rows:", nrow(pos2_70_79), "\n")
fwrite(pos2_70_79, 
       file = "/home/mikej10/advbfx/intronretention/hnRNPH1_IR_MAF/results/splice_out_v2.5/pos2_70_79_filtered.tsv",
       sep = "\t")
cat("  Saved to: pos2_70_79_filtered.tsv\n")

# 2. Extract rows where max1 is less than 0.05
max1_lt_005 <- all_data_valid[max1 < 0.05]
cat("\nExtracting rows where max1 < 0.05:\n")
cat("  Number of rows:", nrow(max1_lt_005), "\n")
fwrite(max1_lt_005,
       file = "/home/mikej10/advbfx/intronretention/hnRNPH1_IR_MAF/results/splice_out_v2.5/max1_lt_005_filtered.tsv",
       sep = "\t")
cat("  Saved to: max1_lt_005_filtered.tsv\n")
```

## 1. Histogram: Position Comparison (pos1 vs pos2)

```{r histogram-pos1-pos2}
# Create histogram comparing pos1 and pos2
pos_data_long <- all_data_valid %>%
  select(decoyID, species, pos1, pos2) %>%
  pivot_longer(cols = c(pos1, pos2), 
               names_to = "position_type", 
               values_to = "position") %>%
  filter(!is.na(position))

p_hist_pos <- ggplot(pos_data_long, aes(x = position, fill = position_type)) +
  geom_histogram(alpha = 0.6, bins = 100, position = "identity") +
  scale_fill_manual(values = c("pos1" = "steelblue", "pos2" = "salmon"),
                    labels = c("pos1", "pos2"),
                    name = "Position Type") +
  scale_x_log10() +
  labs(title = "Distribution of pos1 and pos2 Values (Log Scale)",
       subtitle = "Comparison of canonical splice site position (pos1) vs secondary position (pos2)",
       x = "Position (nucleotides from sequence start, log10 scale)",
       y = "Frequency",
       caption = paste("Total data points:", nrow(pos_data_long))) +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12),
        legend.position = "right") +
  # Add vertical line at position 50 (expected canonical splice site)
  geom_vline(xintercept = 50, linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = 50, y = Inf, label = "Expected pos1 (~50)", 
           vjust = 1.5, hjust = 0, color = "red", fontface = "bold")

print(p_hist_pos)

# Summary statistics
cat("\nPosition Summary Statistics:\n")
cat("  pos1 - Mean:", mean(all_data_valid$pos1, na.rm = TRUE), 
    "| Median:", median(all_data_valid$pos1, na.rm = TRUE), 
    "| % < 55:", round(100 * sum(all_data_valid$pos1 < 55, na.rm = TRUE) / nrow(all_data_valid), 2), "%\n")
cat("  pos2 - Mean:", mean(all_data_valid$pos2, na.rm = TRUE), 
    "| Median:", median(all_data_valid$pos2, na.rm = TRUE),
    "| % < 55:", round(100 * sum(all_data_valid$pos2 < 55, na.rm = TRUE) / nrow(all_data_valid), 2), "%\n")
```

## 2. Scatterplot: max2 > max1 Cases (max2 vs pos2)

```{r scatterplot-max2-gt-max1}
# Filter for cases where max2 > max1
max2_gt_max1_data <- all_data_valid %>%
  filter(max2 > max1) %>%
  filter(!is.na(max2) & !is.na(pos2))

cat("\nCases where max2 > max1:\n")
cat("  Total cases:", nrow(max2_gt_max1_data), "\n")
cat("  Percentage of all valid data:", 
    round(100 * nrow(max2_gt_max1_data) / nrow(all_data_valid), 2), "%\n")

p_scatter_max2 <- ggplot(max2_gt_max1_data, aes(x = pos2, y = max2)) +
  geom_point(alpha = 0.3, size = 0.5, color = "darkgreen") +
  geom_smooth(method = "loess", se = TRUE, color = "red", linewidth = 1.5) +
  scale_x_log10() +
  labs(title = "Scatterplot: max2 vs pos2 (when max2 > max1, Log Scale)",
       subtitle = paste("Cases where secondary score exceeds primary score (n =", 
                       format(nrow(max2_gt_max1_data), big.mark = ","), ")"),
       x = "pos2 (position of max2, log10 scale)",
       y = "max2 (SpliceAI score)",
       caption = "Red line shows LOESS smoothing trend") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12)) +
  scale_y_continuous(limits = c(0, 1)) +
  # Add reference lines
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "blue", alpha = 0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "blue", alpha = 0.5) +
  annotate("text", x = Inf, y = 0.1, label = "max2 = 0.1", 
           hjust = 1.1, vjust = -0.5, color = "blue", size = 3) +
  annotate("text", x = Inf, y = 0.5, label = "max2 = 0.5", 
           hjust = 1.1, vjust = -0.5, color = "blue", size = 3)

print(p_scatter_max2)
```

## 3. Density Plots (Bell Curves) for All Four Statistics

```{r density-plots-all-stats}
# Prepare data for density plots
density_data <- all_data_valid %>%
  select(max1, pos1, max2, pos2) %>%
  pivot_longer(cols = everything(), 
               names_to = "statistic", 
               values_to = "value") %>%
  filter(!is.na(value))

# Create separate plots for pos1/pos2 with log scale and max1/max2 without log scale
density_data_pos <- density_data %>% filter(statistic %in% c("pos1", "pos2"))
density_data_score <- density_data %>% filter(statistic %in% c("max1", "max2"))

p_density_pos <- ggplot(density_data_pos, aes(x = value, fill = statistic)) +
  geom_density(alpha = 0.6, linewidth = 1) +
  facet_wrap(~ statistic, scales = "free", ncol = 2,
             labeller = labeller(statistic = c(
               "pos1" = "pos1 (Position of max1, Log Scale)",
               "pos2" = "pos2 (Position of max2, Log Scale)"
             ))) +
  scale_fill_viridis_d(option = "plasma", name = "Statistic") +
  scale_x_log10() +
  labs(title = "Distribution (Density) of Position Statistics (Log Scale)",
       subtitle = "Bell curves showing the distribution of pos1 and pos2",
       x = "Value (log10 scale)",
       y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12),
        strip.text = element_text(size = 11, face = "bold"),
        legend.position = "none")

p_density_score <- ggplot(density_data_score, aes(x = value, fill = statistic)) +
  geom_density(alpha = 0.6, linewidth = 1) +
  facet_wrap(~ statistic, scales = "free", ncol = 2,
             labeller = labeller(statistic = c(
               "max1" = "max1 (Primary SpliceAI Score)",
               "max2" = "max2 (Secondary SpliceAI Score)"
             ))) +
  scale_fill_viridis_d(option = "plasma", name = "Statistic") +
  labs(title = "Distribution (Density) of Score Statistics",
       subtitle = "Bell curves showing the distribution of max1 and max2",
       x = "Value",
       y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12),
        strip.text = element_text(size = 11, face = "bold"),
        legend.position = "none")

print(p_density_pos)
print(p_density_score)

# Summary statistics for each
cat("\nSummary Statistics by Statistic:\n")
for (stat in c("max1", "pos1", "max2", "pos2")) {
  stat_data <- density_data %>% filter(statistic == stat)
  cat(sprintf("  %s: Mean = %.3f, Median = %.3f, SD = %.3f, Range = [%.3f, %.3f]\n",
              stat,
              mean(stat_data$value, na.rm = TRUE),
              median(stat_data$value, na.rm = TRUE),
              sd(stat_data$value, na.rm = TRUE),
              min(stat_data$value, na.rm = TRUE),
              max(stat_data$value, na.rm = TRUE)))
}
```

## 4. Heatmap of max2 Values

```{r heatmap-max2}
# Load species order from spliceai_heatmap.Rmd approach
# Read species order and mapping
species_order_file <- fread("/home/mikej10/advbfx/intronretention/hnRNPH1_IR_MAF/data/Newickspeciesorder.txt", header = FALSE, col.names = "tree_name")
species_mapping <- fread("/home/mikej10/advbfx/intronretention/hnRNPH1_IR_MAF/results/tree_to_clade_mapping.tsv", sep = "\t", header = TRUE)
ordered_species <- merge(species_order_file, species_mapping, by = "tree_name", all.x = TRUE, sort = FALSE)
ordered_species_ids <- ordered_species$species_id[!is.na(ordered_species$species_id)]

# Map species IDs to common names
species_id_to_common <- setNames(species_mapping$matched_common_name, species_mapping$species_id)

# Prepare matrix for heatmap: decoy IDs as columns, species as rows
# Extract max2 values for each decoy-species pair
max2_matrix_data <- all_data_valid %>%
  select(decoyID, species, max2) %>%
  pivot_wider(names_from = decoyID, values_from = max2)

# Set species as row names and convert to matrix
max2_matrix <- as.matrix(max2_matrix_data[, -1])  # Exclude species column
rownames(max2_matrix) <- max2_matrix_data$species

# Replace NA with -1 for visualization (missing data)
max2_matrix[is.na(max2_matrix)] <- -1

# Get d values for each decoy-species pair (for filtering)
d_matrix_data <- all_data_valid %>%
  select(decoyID, species, d) %>%
  pivot_wider(names_from = decoyID, values_from = d)
d_matrix <- as.matrix(d_matrix_data[, -1])
rownames(d_matrix) <- d_matrix_data$species

# Replace NA in d_matrix with 0 (no length change)
d_matrix[is.na(d_matrix)] <- 0

# Reorder rows (species) to match Newick order from spliceai_heatmap.Rmd
# First, match available species to ordered list
available_species <- rownames(max2_matrix)
ordered_available <- intersect(ordered_species_ids, available_species)
# Add any species not in ordered list at the end
other_species <- setdiff(available_species, ordered_available)
final_species_order <- c(ordered_available, other_species)
max2_matrix <- max2_matrix[final_species_order, , drop = FALSE]

# Map species IDs to common names for row labels
current_rownames <- rownames(max2_matrix)
common_names <- ifelse(current_rownames %in% names(species_id_to_common),
  species_id_to_common[current_rownames],
  current_rownames
)
rownames(max2_matrix) <- common_names

cat("\nMax2 Heatmap Matrix:\n")
cat("  Dimensions:", dim(max2_matrix), "\n")
cat("  Species (rows):", nrow(max2_matrix), "\n")
cat("  Decoy IDs (columns):", ncol(max2_matrix), "\n")
cat("  Value range:", min(max2_matrix), "to", max(max2_matrix), "\n")
cat("  -1 values (missing):", sum(max2_matrix == -1), "\n")

# Sanity check: Verify target decoy exists in matrix
# Try both possible variations of the decoy ID
target_decoy_variants <- c("HNRNPH_1179620582", "HNRNPH1_179620582", "HNRNPH1_1179620582")
target_decoy <- NULL
for (variant in target_decoy_variants) {
  if (variant %in% colnames(max2_matrix)) {
    target_decoy <- variant
    cat("\n✓ Sanity check PASSED: Target decoy '", target_decoy, "' found in matrix\n", sep = "")
    break
  }
}

if (is.null(target_decoy)) {
  cat("\n⚠️  WARNING: Target decoy not found in matrix!\n")
  cat("  Searched for:", paste(target_decoy_variants, collapse = ", "), "\n")
  cat("  First 10 decoy IDs in matrix:", paste(head(colnames(max2_matrix), 10), collapse = ", "), "\n")
  cat("  Checking for partial matches...\n")
  # Check for partial matches
  for (variant in target_decoy_variants) {
    partial_matches <- grep(variant, colnames(max2_matrix), value = TRUE)
    if (length(partial_matches) > 0) {
      cat("  Found partial matches for '", variant, "': ", paste(partial_matches, collapse = ", "), "\n", sep = "")
    }
  }
  # Check for HNRNPH decoys in general
  hnrnph_matches <- grep("^HNRNPH", colnames(max2_matrix), value = TRUE)
  if (length(hnrnph_matches) > 0) {
    cat("  Found ", length(hnrnph_matches), " decoy IDs starting with 'HNRNPH':\n", sep = "")
    cat("    ", paste(head(hnrnph_matches, 20), collapse = ", "), "\n", sep = "")
    if (length(hnrnph_matches) > 20) {
      cat("    ... and ", length(hnrnph_matches) - 20, " more\n", sep = "")
    }
  }
  stop("Target decoy not found. Please verify the decoy ID in the data.")
}

# Create custom color palette with 4 breaks
# Break 1: 0 - 0.1: dark purple to cyan
# Break 2: 0.1 - 0.3: cyan to dark green
# Break 3: 0.3 - 0.7: dark green to bright orange
# Break 4: 0.7 - 1.0: bright orange to hot pink
create_custom_palette <- function(n_colors_per_break = 25) {
  # Break 1: dark purple (#4B0082) to cyan (#00FFFF)
  break1 <- colorRampPalette(c("#4B0082", "#00FFFF"))(n_colors_per_break)
  
  # Break 2: cyan (#00FFFF) to dark green (#006400)
  break2 <- colorRampPalette(c("#00FFFF", "#006400"))(n_colors_per_break)
  
  # Break 3: dark green (#006400) to bright orange (#FF6600)
  break3 <- colorRampPalette(c("#006400", "#FF6600"))(n_colors_per_break)
  
  # Break 4: bright orange (#FF6600) to hot pink (#FF1493)
  break4 <- colorRampPalette(c("#FF6600", "#FF1493"))(n_colors_per_break)
  
  # Combine all breaks
  c("grey70", break1, break2, break3, break4)
}

# Create column annotation for target decoy
annotation_col <- data.frame(
  HNRNPH_target = ifelse(colnames(max2_matrix) == target_decoy, "Yes", "No"),
  row.names = colnames(max2_matrix)
)

# Verify annotation was created correctly
target_count <- sum(annotation_col$HNRNPH_target == "Yes")
cat("  Annotation created: ", target_count, " column(s) marked as target decoy\n", sep = "")

# Create annotation colors
ann_colors <- list(
  HNRNPH_target = c("Yes" = "red", "No" = "white")
)

# Create custom color palette
custom_palette <- create_custom_palette(n_colors_per_break = 25)
n_total_colors <- length(custom_palette) - 1  # Exclude grey

# Create breaks for the color palette
# Breaks: -1.5 to -0.5 (grey), then 0 to 0.1, 0.1 to 0.3, 0.3 to 0.7, 0.7 to 1.0
breaks <- c(-1.5, -0.5, 
            seq(0, 0.1, length.out = 26)[-26],  # 25 colors for 0-0.1
            seq(0.1, 0.3, length.out = 26)[-1],  # 25 colors for 0.1-0.3
            seq(0.3, 0.7, length.out = 26)[-1],  # 25 colors for 0.3-0.7
            seq(0.7, 1.0, length.out = 26)[-1])  # 25 colors for 0.7-1.0

# Create heatmap with species in Newick order (no row clustering)
# Plot full matrix - no sampling (NO d-based filtering in first heatmap)
pheatmap(
  max2_matrix,
  color = custom_palette,
  breaks = breaks,
  cluster_rows = FALSE,  # Keep species in Newick order
  cluster_cols = TRUE,
  show_rownames = TRUE,
  show_colnames = FALSE,  # Too many decoy IDs to show
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  main = "Heatmap of max2 (Secondary SpliceAI Scores) - Species Ordered by Newick Tree",
  fontsize_row = 8,
  fontsize_col = 6,
  border_color = NA,
  na_col = "grey70",
  legend_breaks = c(-1, 0, 0.1, 0.3, 0.7, 1.0),
  legend_labels = c("Missing", "0.0", "0.1", "0.3", "0.7", "1.0")
)
```

## 4b. Heatmap of max1 Values

```{r heatmap-max1}
# Prepare matrix for heatmap: decoy IDs as columns, species as rows
# Extract max1 values for each decoy-species pair
max1_matrix_data <- all_data_valid %>%
  select(decoyID, species, max1) %>%
  pivot_wider(names_from = decoyID, values_from = max1)

# Set species as row names and convert to matrix
max1_matrix <- as.matrix(max1_matrix_data[, -1])  # Exclude species column
rownames(max1_matrix) <- max1_matrix_data$species

# Replace NA with -1 for visualization (missing data)
max1_matrix[is.na(max1_matrix)] <- -1

# Reorder rows (species) to match Newick order (reuse from max2 heatmap)
max1_matrix <- max1_matrix[final_species_order, , drop = FALSE]

# Map species IDs to common names for row labels (reuse from max2 heatmap)
current_rownames_max1 <- rownames(max1_matrix)
common_names_max1 <- ifelse(current_rownames_max1 %in% names(species_id_to_common),
  species_id_to_common[current_rownames_max1],
  current_rownames_max1
)
rownames(max1_matrix) <- common_names_max1

cat("\nMax1 Heatmap Matrix:\n")
cat("  Dimensions:", dim(max1_matrix), "\n")
cat("  Species (rows):", nrow(max1_matrix), "\n")
cat("  Decoy IDs (columns):", ncol(max1_matrix), "\n")
cat("  Value range:", min(max1_matrix), "to", max(max1_matrix), "\n")
cat("  -1 values (missing):", sum(max1_matrix == -1), "\n")

# Use the same custom color palette and breaks as max2
# Create heatmap with species in Newick order (no row clustering)
pheatmap(
  max1_matrix,
  color = custom_palette,
  breaks = breaks,
  cluster_rows = FALSE,  # Keep species in Newick order
  cluster_cols = TRUE,
  show_rownames = TRUE,
  show_colnames = FALSE,  # Too many decoy IDs to show
  main = "Heatmap of max1 (Primary SpliceAI Scores) - Species Ordered by Newick Tree",
  fontsize_row = 8,
  fontsize_col = 6,
  border_color = NA,
  na_col = "grey70",
  legend_breaks = c(-1, 0, 0.1, 0.3, 0.7, 1.0),
  legend_labels = c("Missing", "0.0", "0.1", "0.3", "0.7", "1.0")
)
```

## 5. Scatterplot: max1 < 0.3 Analysis

```{r scatterplot-max1-lt-03}
# Subset data where max1 < 0.3
max1_lt_03_data <- all_data_valid %>%
  filter(max1 < 0.3) %>%
  filter(!is.na(max1) & !is.na(max2))

cat("\nData points with max1 < 0.3:\n")
cat("  Total cases:", nrow(max1_lt_03_data), "\n")
cat("  Percentage of all valid data:", 
    round(100 * nrow(max1_lt_03_data) / nrow(all_data_valid), 2), "%\n")

# Create scatterplot with max1 on y-axis and max2 on x-axis
p_scatter_max1_lt_03 <- ggplot(max1_lt_03_data, aes(x = max2, y = max1)) +
  geom_point(alpha = 0.5, size = 1, color = "steelblue") +
  geom_smooth(method = "loess", se = TRUE, color = "red", linewidth = 1.5) +
  labs(
    title = "Scatterplot: max1 vs max2 (when max1 < 0.3)",
    subtitle = paste("Cases with low primary splice site scores (n =", 
                     format(nrow(max1_lt_03_data), big.mark = ","), ")"),
    x = "max2 (Secondary SpliceAI Score)",
    y = "max1 (Primary SpliceAI Score)",
    caption = "Red line shows LOESS smoothing trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  ) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 0.3)) +
  # Add reference lines
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "blue", alpha = 0.5) +
  geom_hline(yintercept = 0.2, linetype = "dashed", color = "blue", alpha = 0.5) +
  geom_vline(xintercept = 0.3, linetype = "dashed", color = "green", alpha = 0.5) +
  annotate("text", x = Inf, y = 0.1, label = "max1 = 0.1", 
           hjust = 1.1, vjust = -0.5, color = "blue", size = 3) +
  annotate("text", x = Inf, y = 0.2, label = "max1 = 0.2", 
           hjust = 1.1, vjust = -0.5, color = "blue", size = 3) +
  annotate("text", x = 0.3, y = Inf, label = "max2 = 0.3", 
           hjust = 0, vjust = 1.5, color = "green", size = 3)

print(p_scatter_max1_lt_03)

# Summary statistics
cat("\nSummary Statistics for max1 < 0.3:\n")
cat("  Mean max1:", round(mean(max1_lt_03_data$max1, na.rm = TRUE), 4), "\n")
cat("  Mean max2:", round(mean(max1_lt_03_data$max2, na.rm = TRUE), 4), "\n")
cat("  Median max1:", round(median(max1_lt_03_data$max1, na.rm = TRUE), 4), "\n")
cat("  Median max2:", round(median(max1_lt_03_data$max2, na.rm = TRUE), 4), "\n")
cat("  Cases with max2 > 0.3:", sum(max1_lt_03_data$max2 > 0.3, na.rm = TRUE), 
    sprintf("(%.2f%%)\n", 100 * sum(max1_lt_03_data$max2 > 0.3, na.rm = TRUE) / nrow(max1_lt_03_data)))
```

## 6. Extract Aves and Fish Species with max2 > 0.3

```{r extract-aves-fish-max2-gt-03}
# Identify aves and fish species from tree_to_clade_mapping
# (species_mapping already loaded in heatmap section)
aves_fish_species <- species_mapping %>%
  filter(clade %in% c("Aves", "Fish")) %>%
  select(species_id, clade, matched_common_name)

cat("\nAves and Fish Species Identification:\n")
cat("  Total Aves species:", sum(aves_fish_species$clade == "Aves"), "\n")
cat("  Total Fish species:", sum(aves_fish_species$clade == "Fish"), "\n")
cat("  Total Aves + Fish species:", nrow(aves_fish_species), "\n")

# Get list of species IDs for aves and fish
aves_fish_species_ids <- aves_fish_species$species_id

# Subset data where max2 > 0.3 in aves or fish species
aves_fish_max2_gt_03 <- all_data_valid %>%
  filter(species %in% aves_fish_species_ids) %>%
  filter(max2 > 0.3) %>%
  filter(!is.na(max2))

# Add clade and common name information
aves_fish_max2_gt_03 <- aves_fish_max2_gt_03 %>%
  left_join(aves_fish_species, by = c("species" = "species_id"))

cat("\nData points with max2 > 0.3 in Aves or Fish species:\n")
cat("  Total cases:", nrow(aves_fish_max2_gt_03), "\n")
cat("  Percentage of all valid data:", 
    round(100 * nrow(aves_fish_max2_gt_03) / nrow(all_data_valid), 2), "%\n")
cat("  Breakdown by clade:\n")
clade_breakdown <- aves_fish_max2_gt_03 %>%
  count(clade) %>%
  mutate(percentage = round(100 * n / sum(n), 2))
print(clade_breakdown)

# Save to TSV file
output_file <- "/home/mikej10/advbfx/intronretention/hnRNPH1_IR_MAF/results/splice_out_v2.5/aves_fish_max2_gt_03.tsv"
fwrite(aves_fish_max2_gt_03, file = output_file, sep = "\t")
cat("\n  Saved to: aves_fish_max2_gt_03.tsv\n")

# Summary statistics
cat("\nSummary Statistics for Aves/Fish with max2 > 0.3:\n")
cat("  Mean max1:", round(mean(aves_fish_max2_gt_03$max1, na.rm = TRUE), 4), "\n")
cat("  Mean max2:", round(mean(aves_fish_max2_gt_03$max2, na.rm = TRUE), 4), "\n")
cat("  Median max1:", round(median(aves_fish_max2_gt_03$max1, na.rm = TRUE), 4), "\n")
cat("  Median max2:", round(median(aves_fish_max2_gt_03$max2, na.rm = TRUE), 4), "\n")
cat("  Unique decoy IDs:", length(unique(aves_fish_max2_gt_03$decoyID)), "\n")
cat("  Unique species:", length(unique(aves_fish_max2_gt_03$species)), "\n")

# Show top species by count
cat("\nTop 10 species by occurrence count:\n")
top_species <- aves_fish_max2_gt_03 %>%
  count(matched_common_name, clade, sort = TRUE) %>%
  head(10)
print(top_species)
```

## 7. Summary: Validation of Canonical Splice Site Detection

```{r validation-summary}
cat("\n")
cat(paste0(rep("=", 80), collapse = ""), "\n")
cat("VALIDATION SUMMARY: Canonical Splice Site Detection\n")
cat(paste0(rep("=", 80), collapse = ""), "\n\n")

# Key metrics
total_valid <- nrow(all_data_valid)
pos1_near_50 <- sum(all_data_valid$pos1 >= 40 & all_data_valid$pos1 <= 60, na.rm = TRUE)
pos1_lt_55 <- sum(all_data_valid$pos1 < 55, na.rm = TRUE)
pos2_lt_55 <- sum(all_data_valid$pos2 < 55, na.rm = TRUE)
max1_high <- sum(all_data_valid$max1 > 0.8, na.rm = TRUE)
max2_gt_max1 <- sum(all_data_valid$max2 > all_data_valid$max1, na.rm = TRUE)
max2_gt_07 <- sum(all_data_valid$max2 > 0.7, na.rm = TRUE)
max2_gt_08 <- sum(all_data_valid$max2 > 0.8, na.rm = TRUE)
max1_lt_005 <- sum(all_data_valid$max1 < 0.05, na.rm = TRUE)
d_gt_09 <- sum(all_data_valid$d > 0.9, na.rm = TRUE)

cat("Key Validation Metrics:\n")
cat(sprintf("  1. pos1 in expected range (40-60 nt): %d (%.2f%%)\n", 
            pos1_near_50, 100 * pos1_near_50 / total_valid))
cat(sprintf("  2. pos1 < 55 nt: %d (%.2f%%)\n", 
            pos1_lt_55, 100 * pos1_lt_55 / total_valid))
cat(sprintf("  3. pos2 < 55 nt: %d (%.2f%%)\n", 
            pos2_lt_55, 100 * pos2_lt_55 / total_valid))
cat(sprintf("  4. max1 > 0.8 (high confidence): %d (%.2f%%)\n", 
            max1_high, 100 * max1_high / total_valid))
cat(sprintf("  5. max2 > max1 (secondary site stronger): %d (%.2f%%)\n", 
            max2_gt_max1, 100 * max2_gt_max1 / total_valid))
cat(sprintf("  6. max2 > 0.7: %d (%.2f%%)\n", 
            max2_gt_07, 100 * max2_gt_07 / total_valid))
cat(sprintf("  7. max2 > 0.8: %d (%.2f%%)\n", 
            max2_gt_08, 100 * max2_gt_08 / total_valid))
cat(sprintf("  8. max1 < 0.05: %d (%.2f%%)\n", 
            max1_lt_005, 100 * max1_lt_005 / total_valid))
cat(sprintf("  9. d > 0.9: %d (%.2f%%)\n", 
            d_gt_09, 100 * d_gt_09 / total_valid))
cat(sprintf("  10. Mean pos1: %.2f nt (expected ~50 nt)\n", 
            mean(all_data_valid$pos1, na.rm = TRUE)))
cat(sprintf("  11. Median pos1: %.2f nt\n", 
            median(all_data_valid$pos1, na.rm = TRUE)))

# Investigate contradiction: pos1 < 55 should include pos1 in range 40-60
cat("\nInvestigating pos1 range contradiction:\n")
pos1_lt_40 <- sum(all_data_valid$pos1 < 40, na.rm = TRUE)
pos1_40_54 <- sum(all_data_valid$pos1 >= 40 & all_data_valid$pos1 < 55, na.rm = TRUE)
pos1_55_60 <- sum(all_data_valid$pos1 >= 55 & all_data_valid$pos1 <= 60, na.rm = TRUE)
pos1_gt_60 <- sum(all_data_valid$pos1 > 60, na.rm = TRUE)

cat(sprintf("  pos1 < 40: %d (%.2f%%)\n", pos1_lt_40, 100 * pos1_lt_40 / total_valid))
cat(sprintf("  pos1 40-54: %d (%.2f%%)\n", pos1_40_54, 100 * pos1_40_54 / total_valid))
cat(sprintf("  pos1 55-60: %d (%.2f%%)\n", pos1_55_60, 100 * pos1_55_60 / total_valid))
cat(sprintf("  pos1 > 60: %d (%.2f%%)\n", pos1_gt_60, 100 * pos1_gt_60 / total_valid))
cat(sprintf("  Sum check: %d (should equal total_valid: %d)\n", 
            pos1_lt_40 + pos1_40_54 + pos1_55_60 + pos1_gt_60, total_valid))
cat(sprintf("  pos1 < 55 breakdown: < 40 (%d) + 40-54 (%d) = %d\n", 
            pos1_lt_40, pos1_40_54, pos1_lt_40 + pos1_40_54))
cat(sprintf("  pos1 40-60 breakdown: 40-54 (%d) + 55-60 (%d) = %d\n", 
            pos1_40_54, pos1_55_60, pos1_40_54 + pos1_55_60))

cat("\nInterpretation:\n")
if (mean(all_data_valid$pos1, na.rm = TRUE) > 40 && mean(all_data_valid$pos1, na.rm = TRUE) < 60) {
  cat("  ✓ Mean pos1 is in expected range (40-60 nt) - GOOD\n")
} else {
  cat("  ⚠ Mean pos1 is outside expected range - may indicate issues with large insertions\n")
}

if (100 * pos1_lt_55 / total_valid > 80) {
  cat("  ✓ >80% of pos1 values are < 55 nt - GOOD\n")
} else {
  cat("  ⚠ <80% of pos1 values are < 55 nt - may indicate detection issues\n")
}

cat("\n", paste0(rep("=", 80), collapse = ""), "\n")
```
